### 2024-09-24 paper/paper.qmd
Made a new graph

---
title: "Trends in Mortality Rates"
subtitle: "Analyzing Deaths Among Shelter Residents from 2007 to 2024"
author: 
  - Mandy He
thanks: "Code and data are available at: https://open.toronto.ca/dataset/deaths-of-shelter-residents/"
date: Septmeber 23 2024
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
library(ggplot2)
library(dplyr)

Death_data <- read_csv(here::here("data/analysis_data/Num_death_analysis_data.csv"))
AvgAge_data <- read_csv(here::here("data/analysis_data/AvgAge_death_analysis_data.csv"))

```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-Total Number of Death By Year ), from @palmerpenguins.

```{r}
#| label: fig-Total Number of Death By Year 
#| fig-cap: Yearly Mortality rate from 2007-2024
#| echo: false

AvgAge_data |> 
  ggplot(aes(x = year, y = total_decedents)) +
  geom_point()+
  labs(title = "Total Decedents by Year", x = "Year", y = "Total Decedents") +
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5)) + 
  scale_x_continuous(breaks = unique(AvgAge_data$year))+
  scale_y_continuous(breaks = seq(0, max(AvgAge_data$total_decedents), by = 5))

```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-09-25 paper/paper.qmd
add reference and more information

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "Poverty and homelessness remain pressing issues in today’s society, particularly evident in downtown Toronto, where many individuals still sleep on the streets or in subway stations.
+Due to these harsh living conditions, these individuals often face severe health issues, leading to an increased risk of mortality.
+Luckly, there are services in the city that provide place for some of these people to live.
+This paper analyzes the trends in mortality rates among shelter residents in the City of Toronto from 2007 to 2024, highlighting patterns over time and exploring the factors contributing to these rates."
 format: pdf
@@
 bibliography: references.bib
+
 ---
@@
 library(tidyverse)
-library(palmerpenguins)
 library(ggplot2)
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+Homelessness is a well-known issue worldwide, particularly in impoverished cities and countries.
+It was reported approximately 15 million homeless individuals globally by the World Economic Forum in 2021 (@homelessfact).
+Although homelessness is more severe in developing countries, it does not mean that developed places are free from this problem.
+In downtown Toronto, a city that hosts numerous corporate headquarters and some of the Canada's top educational institutions, homelessness remains a significant issue.
+Homeless individuals can be seen on the streets and even in subway stations, showing the persistent challenges faced by the city and those without stable housing.
 
-The remainder of this paper is structured as follows. @sec-data....
+Homeless individuals often live and sleep in harsh and unsanitary conditions, which severely compromises their health.
+They are more likely to get diseases and suffer from various health problems due to limited access to proper healthcare and nutrition.
+Consequently, this significantly increases their risk of mortality.
+The City of Toronto is also deeply concerned about this issue and has implemented numerous aid programs and support services to assist homeless individuals.
+Despite these efforts, many homeless individuals continue to face significant challenges, and there are still a considerable number of deaths occurring at relatively young ages
+
+In this paper we will look at the data @toronto_deaths_shelter_residents, which explore the trends in mortality rates among shelter residents in the City of Toronto from 2007 to 2024, highlighting patterns over time and exploring the factors contributing to these rates.The data indicates that mortality rates were relatively low, with fewer than five deaths per month in the prior 2019.
+However, after 2019, there was a sharp increase in the number of deaths, peaking in 2021 and 2022.
+This surge can be attributed to the onset of the COVID-19 pandemic in 2019, which significantly impacted public health and resulted in many fatalities.
+
+Additionally, the challenging living conditions faced by homeless individuals, compared to those of wealthier individuals who also experienced hardships during this time, suggest that the mortality rates among shelter residents are likely to remain elevated.
+COVID-19 was also a primary driver behind the large number of deaths, as evidenced by the immediate decrease in the number of deaths following the post-pandemic period shown in the graphs.
+The data was examined more in depth by analyzing the mortality rate and average mortality rate between male and female homeless individuals in shelter residences.
+There are a lot more male death compare to female death, but this make sense as the proportion of male residents is greater than that of female residents, with 60% male and only 40% female (@toronto_shelter_flow_data).
+Additionally, the average age of deceased individuals is around 50 to 60 years old (@datacommons_life_expectancy).
+According to data, the life expectancy in Canada is approximately 81 years, indicating that the average age at death for homeless individuals is quite young.
+
+The remainder of this paper is structured as follows.
+Section 2 will discuss the data used in the study, highlighting key aspects and limitations.
+Section 3 will focus on generating graphs using the data presented in Section 2.
+Section 4 will elaborate on the graphs from Section 3, providing interpretations and possible explanations for the findings.
 
@@
 
-Some of our data is of penguins (@fig-Total Number of Death By Year ), from @palmerpenguins.
+## Raw Data {#sec-first-point}
+
+The data used in this paper is access in from Open Data Toronto and the particular data set used was the Deaths of Shelter Residents (@toronto_deaths_shelter_residents).
+To analysis the data and creating graphs using the data, following package that was build in the R program @citeR was used: tidyverse (@tidyverse), dplyr (@dplyr), knitr (@knitr), and ggplot2 ()@ggplot2.
+
+The data used in this paper was collected by the Toronto Shelter and Support Services Division since 2007.
+Shelters are required to notify the City of Toronto within 24 hours of a death and submit a written report within 30 days.
+Consequently, the data is updated monthly on the Open Data Toronto website.
+The raw data set includes 10 variables and 18 observations, covering the period from 2007 to 2024, as shown in Table 1 (@Tab-Table1).
+The key variables used in this paper include: year, total decedents, average age at death for all decedents, male decedents, average age at death for male decedents, female decedents, and average age at death for female decedents.
+Additionally, the monthly death counts from 2007 to 2024 will be used for graphing, comprising 212 observations.
 
 ```{r}
-#| label: fig-Total Number of Death By Year 
+#| label: Tab-Table1
+#| fig-cap: Yearly Mortality rate from 2007-2024
+#| echo: false
+
+library(knitr)
+library(dplyr)
+
+# Remove the 'id' column
+AvgAge_data_no_id <- AvgAge_data %>%
+  select(-id)
+
+# Display the table without the 'id' column
+kable(AvgAge_data_no_id, 
+      caption = "Trends in Mortality Rates Among Shelter Residents (2007-2024)",
+      col.names = c("Year", "Total Decedents", 
+                    "Avg Age of Death (All)", "Male Decedents", 
+                    "Avg Age of Death (Male)", "Female Decedents", 
+                    "Avg Age of Death (Female)", "Trans/NB/2S Decedents", 
+                    "Avg Age of Death (Trans/NB/2S)"))
+```
+
+## Limitation of Data {#sec-first-point}
+
+The Deaths of Shelter Residents data is part of a larger database shared by Toronto Public Health (TPH), which also includes deaths of homeless individuals living in shelters and outdoors.
+Therefore, the findings in this paper focus solely on homeless individuals residing in shelters, which may limit the dataset and lead to potential biases that do not fully represent the mortality of all homeless individuals.
+
+Other limitations include the absence of certain data points, such as the cause of death and additional information that could identify the deceased, due to confidentiality concerns.
+This lack of information may impact the reasoning behind the findings, as possible explanations will need to be drawn from other research and sources.
+
+# Results
+
+@fig-TotalNumberofDeathByYear is showing the total amount of decedent of homeless living in the Shelter.
+
+```{r}
+#| label: fig-TotalNumberofDeathByYear 
 #| fig-cap: Yearly Mortality rate from 2007-2024
@@
   ggplot(aes(x = year, y = total_decedents)) +
-  geom_point()+
+  geom_point() +
+  geom_line() +
   labs(title = "Total Decedents by Year", x = "Year", y = "Total Decedents") +
-  theme_minimal()+
+  theme_minimal() +
   theme(plot.title = element_text(hjust = 0.5)) + 
-  scale_x_continuous(breaks = unique(AvgAge_data$year))+
-  scale_y_continuous(breaks = seq(0, max(AvgAge_data$total_decedents), by = 5))
+  scale_x_continuous(breaks = unique(AvgAge_data$year)) +
+  scale_y_continuous(breaks = seq(0, max(AvgAge_data$total_decedents), by = 20))
 
@@
 
-Talk more about it.
+@fig-TotalDecedentsbyMonth shows the number of deaths in each month-year pair from 2007 to 2024. 
+```{r}
+#| label: fig-TotalDecedentsbyMonth
+#| fig-cap: Yearly Mortality rate from 2007-2024
+#| echo: false
 
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
+# Combine Year and Month into a single column with last two digits of the year
+Death_data$YearMonth <- paste(substr(Death_data$year, 3, 4), Death_data$month, sep = "-")
 
+# Convert YearMonth to a factor with ordered levels
+Death_data$YearMonth <- factor(Death_data$YearMonth, levels = unique(Death_data$YearMonth))
+
+# Plot the graph
+Death_data |>
+  ggplot(aes(x = YearMonth, y = total_decedents, group=1)) +
+  geom_line(color="blue") +
+  geom_point(color="red") +
+  scale_x_discrete(breaks = Death_data$YearMonth[seq(1, nrow(Death_data), by = 12)]) + # Show only every 12th month (i.e., January each year)
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) + # Display x-axis labels vertically
+  labs(title="Trends in Total Decedents by Month and Year", 
+       x="Month-Year", 
+       y="Total Decedents") +
+  theme_minimal()
+```
+
+Talk way more about it.
+
+@fig-TotalDecedentbyMonthcovid is a close up look from @fig-TotalDecedentsbyMonth focusing on data ranging from year 2019 to 2022, during the pandemic. 
 ```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
+#| label: fig-TotalDecedentbyMonthcovid
+#| fig-cap: Yearly Mortality rate from 2019-2022
+#| echo: false
+
+# Filter the data for years 2019 to 2022
+Death_data_filtered <- Death_data %>%
+  filter(year >= 2019 & year <= 2022)
+
+# Combine Year and Month into a single column with last two digits of the year
+Death_data_filtered$YearMonth <- paste(substr(Death_data_filtered$year, 3, 4), Death_data_filtered$month, sep = "-")
+
+# Convert YearMonth to a factor with ordered levels
+Death_data_filtered$YearMonth <- factor(Death_data_filtered$YearMonth, levels = unique(Death_data_filtered$YearMonth))
+
+# Plot the graph for filtered data
+Death_data_filtered |>
+  ggplot(aes(x = YearMonth, y = total_decedents, group = 1)) +
+  geom_line(color = "blue") +
+  geom_point(color = "red") +
+  scale_x_discrete(breaks = Death_data_filtered$YearMonth[seq(1, nrow(Death_data_filtered), by = 6)]) + # Show every 6th month
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) + # Display x-axis labels vertically
+  labs(title = "Trends in Total Decedents by Month and Year (2019-2022)",
+       x = "Month-Year",
+       y = "Total Decedents") +
+  theme_minimal()
+```
+
+@ fig-TotalDecedentsFemaleVsMale display a comparison in number of death between females and males.
+```{r}
+#| label: fig-TotalDecedentsFemaleVsMale 
+#| fig-cap: Total Decedents Female VS Male
+#| echo: false
+#| include: false
+#| warning: false
+#| message: false
+
+library(ggplot2)
+library(tidyr)
+library(dplyr)
+
+# Reshape the data for plotting
+plot_data <- AvgAge_data %>%
+  select(year, Male_decedents, Female_decedents) %>%
+  pivot_longer(cols = c(Male_decedents, Female_decedents),
+               names_to = "Gender",
+               values_to = "Mortality")
+
+# Create a new variable for the last two digits of the year with leading zero
+plot_data <- plot_data %>%
+  mutate(year_short = sprintf("%02d", year %% 100))
+
+# Create the bar graph with adjusted year labels
+ggplot(plot_data, aes(x = year_short, y = Mortality, fill = Gender)) +
+  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
+  labs(title = "Mortality Numbers of Male and Female Decedents",
+       x = "Year (2000s)",
+       y = "Number of Decedents") +
+  scale_fill_manual(values = c("Male_decedents" = "blue", "Female_decedents" = "pink")) +
+  theme_minimal()
+
+```
+
+@fig-AvgAgeofDecedentsFemaleVSMale presents a comparison of the average age at death between females and males.
+```{r}
+#| label: fig-AvgAgeofDecedentsFemaleVSMale 
+#| fig-cap: Total Decedents Female VS Male
 #| echo: false
@@
 
-analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
+# Reshape the data for plotting average age of death
+avg_age_data <- AvgAge_data %>%
+  select(year, Average_age_of_death_male_decedents, Average_age_of_death_female_decedents) %>%
+  pivot_longer(cols = c(Average_age_of_death_male_decedents, Average_age_of_death_female_decedents),
+               names_to = "Gender",
+               values_to = "Average_Age") %>%
+  mutate(year_short = sprintf("%02d", year %% 100))
 
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
+# Create the line graph for average age of death
+ggplot(avg_age_data, aes(x = year_short, y = Average_Age, color = Gender, group = Gender)) +
+  geom_line(size = 1) +   
+  geom_point(size = 2) +    
+  labs(title = "Average Age of Death for Male and Female Decedents",
+       x = "Year",
+       y = "Average Age of Death") +
+  scale_color_manual(values = c("Average_age_of_death_male_decedents" = "blue", 
+                                 "Average_age_of_death_female_decedents" = "pink")) +
   theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
-```
+  theme(legend.title = element_blank(),
+        legend.position = "bottom")
 
-Talk way more about it.
-
-# Results
-
-Our results are summarized in @tbl-modelresults.
-
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
-```
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
 ```
@@
 
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
-
 ## Diagnostics
@@
 
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
-```
-
 \newpage

### 2024-09-25 paper/paper.qmd
new info added

@@
 
-Additionally, the challenging living conditions faced by homeless individuals, compared to those of wealthier individuals who also experienced hardships during this time, suggest that the mortality rates among shelter residents are likely to remain elevated.
-COVID-19 was also a primary driver behind the large number of deaths, as evidenced by the immediate decrease in the number of deaths following the post-pandemic period shown in the graphs.
 The data was examined more in depth by analyzing the mortality rate and average mortality rate between male and female homeless individuals in shelter residences.
-There are a lot more male death compare to female death, but this make sense as the proportion of male residents is greater than that of female residents, with 60% male and only 40% female (@toronto_shelter_flow_data).
-Additionally, the average age of deceased individuals is around 50 to 60 years old (@datacommons_life_expectancy).
-According to data, the life expectancy in Canada is approximately 81 years, indicating that the average age at death for homeless individuals is quite young.
+There are a lot more male death compare to female death every single year.
+Additionally, the average age at death is around 50 to 60 years old, which is relatively young compared to Canada’s life expectancy.
+Future details about the findings will be discussed in the discussion section.
 
@@
 The data used in this paper is access in from Open Data Toronto and the particular data set used was the Deaths of Shelter Residents (@toronto_deaths_shelter_residents).
-To analysis the data and creating graphs using the data, following package that was build in the R program @citeR was used: tidyverse (@tidyverse), dplyr (@dplyr), knitr (@knitr), and ggplot2 ()@ggplot2.
+To analysis the data and creating graphs using the data, following package that was build in the (R program @citeR) was used: tidyverse (@tidyverse), dplyr (@dplyr), knitr (@knitr), and ggplot2 (@ggplot2).
 
@@
 
-The Deaths of Shelter Residents data is part of a larger database shared by Toronto Public Health (TPH), which also includes deaths of homeless individuals living in shelters and outdoors.
-Therefore, the findings in this paper focus solely on homeless individuals residing in shelters, which may limit the dataset and lead to potential biases that do not fully represent the mortality of all homeless individuals.
+The Deaths of Shelter Residents data is part of a larger database shared by Toronto Public Health (TPH), which also includes deaths of homeless individuals living in shelters and outdoors (@toronto_homeless_deaths).
+Therefore, the findings in this paper focus solely on homeless individuals residing in shelters, which may limit the data set and lead to potential biases that do not fully represent the mortality of all homeless individuals.
 
@@
 
+\newpage
+
 # Results
-
-@fig-TotalNumberofDeathByYear is showing the total amount of decedent of homeless living in the Shelter.
 
@@
 
-@fig-TotalDecedentsbyMonth shows the number of deaths in each month-year pair from 2007 to 2024. 
 ```{r}
@@
 
-Talk way more about it.
-
-@fig-TotalDecedentbyMonthcovid is a close up look from @fig-TotalDecedentsbyMonth focusing on data ranging from year 2019 to 2022, during the pandemic. 
 ```{r}
 #| label: fig-TotalDecedentbyMonthcovid
-#| fig-cap: Yearly Mortality rate from 2019-2022
+#| fig-cap: Yearly Mortality rate from 2019-2023
 #| echo: false
@@
 Death_data_filtered <- Death_data %>%
-  filter(year >= 2019 & year <= 2022)
+  filter(year >= 2019 & year <= 2023)
 
@@
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) + # Display x-axis labels vertically
-  labs(title = "Trends in Total Decedents by Month and Year (2019-2022)",
+  labs(title = "Trends in Total Decedents by Month and Year (2019-2023)",
        x = "Month-Year",
@@
 
-@ fig-TotalDecedentsFemaleVsMale display a comparison in number of death between females and males.
 ```{r}
-#| label: fig-TotalDecedentsFemaleVsMale 
+#| label: fig-T 
 #| fig-cap: Total Decedents Female VS Male
 #| echo: false
-#| include: false
 #| warning: false
@@
 
-@fig-AvgAgeofDecedentsFemaleVSMale presents a comparison of the average age at death between females and males.
 ```{r}
@@
 
+\newpage
+
 # Discussion
@@
 ## First discussion point {#sec-first-point}
+@fig-TotalNumberofDeathByYear shows the total number of decedents among homeless individuals living in shelters.
+The graph visualizes two notable increases in 2013 and 2019.
+According to making connection with other sources, the primary causes of these increases are linked to macro health conditions occurring during those times.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+In 2013, there was an influenza outbreak across Ontario that lasted from 2013 to 2016.
+Influenza is a viral infection that affects the nose, throat, and sometimes the lungs, leading to respiratory illness.
+Symptoms can range from mild to severe, and in some cases, the flu can result in death.
+The best way to prevent infection is to get vaccinated each year (@cdc_flu_key_facts).
+There is a clear trend of increasing deaths up until 2015, followed by a drop in 2016.
+This pattern closely resembles the total number of influenza outbreaks reported in the article "Influenza Outbreaks in Ontario Hospitals" (@cameron2019).
+According to Table 1 in the article, the number of outbreaks increased from 38 in 2013-2014 to 117 in 2014-2015, then decreased to 36 in 2015-2016.
+This rise and fall in outbreaks corresponds with the trends observed in @fig-TotalNumberofDeathByYear, where the number of deaths peaked in 2015 and then significantly declined in 2016, coinciding with the flu being brought under control.
+
+One notable point is that in the article, the fatality rate for the 2014-2015 outbreak was only 2.4%, despite the hundreds of flu outbreaks that occurred that year.
+However,@fig-TotalNumberofDeathByYear shows that the number of deaths in 2015 was around 45.
+This discrepancy can be explained by the fact that the data in the article pertains to hospital outbreaks, where individuals have access to better medical care.
+In contrast, homeless individuals may lack the resources to seek treatment, leading to a higher number of deaths in this population.
 
@@
 
+Another clear increasing trend in the mortality rate is evident in 2019, as shown in @fig-TotalNumberofDeathByYear.
+This is not surprising, given that people around the world were experiencing and still recovering from the COVID-19 pandemic, which began in 2019 and continued until the end of 2023.
+To better understand the impact, the Government of Canada (@canada_health_infobase) continuously updates the numbers of cases and deaths related to the pandemic, showing a significant rise in fatalities from 2021 to 2022, followed by a decline after 2023.
+
+Additionally, the challenging living conditions faced by homeless individuals, compared to those of wealthier individuals who also experienced hardships during this time, suggest that the mortality rates among shelter residents are likely to remain elevated.
+COVID-19 was also a primary driver behind the large number of deaths, as evidenced by the immediate decrease in the number of deaths following the post-pandemic period shown in @fig-TotalDecedentbyMonthcovid.
+
 ## Third discussion point
+
+From @fig-T, it is evident that the mortality rate for males is significantly higher than that for females.
+One reason for this observation is that the proportion of male residents in shelters is greater than that of female residents, with 60% male and only 40% female (@toronto_shelter_flow_data).
+Nonetheless, research has shown that females tend to live longer than males at all stages of life, from infancy to young adulthood and into retirement (@ourworldindata_life_expectancy).
+The overall reason for the higher male mortality rate compared to females can be attributed to genetic factors as well as differences in health awareness and behaviors between the two genders.
+
+@fig-AvgAgeofDecedentsFemaleVSMale indicates that the average age at death for homeless individuals living in shelters ranges from 50 to 60 years.
+According to data from the World Bank, the life expectancy in Canada is approximately 81 years, which suggests that the average age at death for homeless individuals is relatively young (@datacommons_life_expectancy).
+Despite the increasing life expectancy in today’s society, as health and technology continue to evolve, homeless individuals still die at a younger age due to financial constraints and limited access to resources.
 
@@
 
-Weaknesses and next steps should also be included.
+The findings in this paper represent a small aspect of what is happening in the world, focusing specifically on the homeless population.
+While these results cannot be used to draw definitive conclusions about broader trends, they do provide insights into patterns that may be occurring on a larger scale.
+
+The connections discussed in this section are not statistically proven, as no models were run to demonstrate the correlation between the deaths of homeless individuals residing in shelters and the various possible explanations mentioned.
+In addition, the actual causes of death may remain unknown due to confidentiality restrictions and are inferred based on global events that could have contributed to these fatalities.
+
+Other data will have to be gathered and analyzed using appropriate models to establish a solid scientific connection.
+For instance, to establish a connection between the increase in death numbers in 2013 and 2019 and the flu outbreak and pandemic, data on flu-related deaths and pandemic-related deaths should be gathered and analyzed.
+A model should be run to test the relationship between these variables.
 

### 2024-09-27 paper/paper.qmd
Added more to discussion section and reformat the graphs

@@
 Due to these harsh living conditions, these individuals often face severe health issues, leading to an increased risk of mortality.
-Luckly, there are services in the city that provide place for some of these people to live.
+Luckily, there are services in the city that provide place for some of these people to live.
 This paper analyzes the trends in mortality rates among shelter residents in the City of Toronto from 2007 to 2024, highlighting patterns over time and exploring the factors contributing to these rates."
@@
 bibliography: references.bib 
+toc: true
 
@@
 The City of Toronto is also deeply concerned about this issue and has implemented numerous aid programs and support services to assist homeless individuals.
-Despite these efforts, many homeless individuals continue to face significant challenges, and there are still a considerable number of deaths occurring at relatively young ages
+Despite these efforts, many homeless individuals continue to face significant challenges, and there are still a considerable number of deaths occurring at relatively young ages.
 
@@
 The remainder of this paper is structured as follows.
-Section 2 will discuss the data used in the study, highlighting key aspects and limitations.
-Section 3 will focus on generating graphs using the data presented in Section 2.
-Section 4 will elaborate on the graphs from Section 3, providing interpretations and possible explanations for the findings.
+@sec-data will discuss the data used in the study, highlighting key aspects and limitations.
+@sec-Results will focus on generating graphs using the data presented in @sec-data.
+@sec-discussion will elaborate on the graphs from @sec-Results, providing interpretations and possible explanations for the findings.
 
@@
 To analysis the data and creating graphs using the data, following package that was build in the (R program @citeR) was used: tidyverse (@tidyverse), dplyr (@dplyr), knitr (@knitr), and ggplot2 (@ggplot2).
+LLM (@openai2023chatgpt) was also use for this paper in term of helping with coding in R to graph data, simulating a data, cleaning data, and revising writing.
 
@@
 Consequently, the data is updated monthly on the Open Data Toronto website.
-The raw data set includes 10 variables and 18 observations, covering the period from 2007 to 2024, as shown in Table 1 (@Tab-Table1).
+The raw data set includes 10 variables and 18 observations, covering the period from 2007 to 2024, as shown in @tbl-Table1.
 The key variables used in this paper include: year, total decedents, average age at death for all decedents, male decedents, average age at death for male decedents, female decedents, and average age at death for female decedents.
@@
 
+**Variable in the data are as following:**
+
+**id:** A unique row identifier is used in the Open Data database, as names are not reported due to confidentiality.
+
+**Year:** The calendar year during which the number of deaths is reported.
+
+**Month:** The month during which the number of deaths is reported.
+
+**Total decedents:** Total number of shelter residents who died in the reported month/year.
+
+**Male:** Total number of male shelter residents who died in the reported month/year.
+
+**Female:** Total number of female shelter residents who died in the reported month/year.
+
+**Transgender/Non-binary/Two-Spirit:** Total number of transgender, non-binary, and Two-Spirit shelter residents who died in the reported month/year.
+
+\newpage
 ```{r}
-#| label: Tab-Table1
-#| fig-cap: Yearly Mortality rate from 2007-2024
+#| label: tbl-Table1
+#| tbl-cap: Trends in Mortality Rates Among Shelter Residents (2007-2024)
 #| echo: false
@@
 kable(AvgAge_data_no_id, 
-      caption = "Trends in Mortality Rates Among Shelter Residents (2007-2024)",
       col.names = c("Year", "Total Decedents", 
@@
 
-## Limitation of Data {#sec-first-point}
+## Limitation of Data {#sec-lim}
 
@@
 
-# Results
+# Results {#sec-Results}
+
+The following graph (@fig-TotalNumberofDeathByYear) shows the overall trend in the number of decedents living in shelters from 2007 to 2024.
+It reveals that between 2007 and 2013, the number of deaths remained quite low, under 30 each year.
+However, starting in 2013, the number of deaths increased, and the mortality rate never returned to the low levels seen from 2007 to 2013.
+The graph also highlights two significant peaks in 2015 and 2021.
+The increase in deaths was more pronounced in 2021 compared to 2015.
+Further discussion on the reasons for these peaks can be found in discussion sections (@sec-first-dpoint and @sec-second-dpoint).
+Furthermore, in the appendix (@sec-app), there are graph for number of death per month for each individual year.
 
@@
   geom_line() +
-  labs(title = "Total Decedents by Year", x = "Year", y = "Total Decedents") +
+  labs(title = "Total Shelter Residents Decedents from 2007 to 2024", x = "Year", y = "Total Decedents") +
   theme_minimal() +
+   theme(axis.text.x = element_text(angle = 45)) +
   theme(plot.title = element_text(hjust = 0.5)) + 
@@
 
-```{r}
-#| label: fig-TotalDecedentsbyMonth
-#| fig-cap: Yearly Mortality rate from 2007-2024
-#| echo: false
-
-# Combine Year and Month into a single column with last two digits of the year
-Death_data$YearMonth <- paste(substr(Death_data$year, 3, 4), Death_data$month, sep = "-")
-
-# Convert YearMonth to a factor with ordered levels
-Death_data$YearMonth <- factor(Death_data$YearMonth, levels = unique(Death_data$YearMonth))
-
-# Plot the graph
-Death_data |>
-  ggplot(aes(x = YearMonth, y = total_decedents, group=1)) +
-  geom_line(color="blue") +
-  geom_point(color="red") +
-  scale_x_discrete(breaks = Death_data$YearMonth[seq(1, nrow(Death_data), by = 12)]) + # Show only every 12th month (i.e., January each year)
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) + # Display x-axis labels vertically
-  labs(title="Trends in Total Decedents by Month and Year", 
-       x="Month-Year", 
-       y="Total Decedents") +
-  theme_minimal()
-```
-
+\newpage
+@fig-TotalDecedentbyMonthcovid provides a close-up view of the years 2019 to 2024, covering the period of the COVID-19 pandemic and a bit of the post-pandemic phase.
+During this time, the number of deaths does not exhibit a clear trend as time progresses.
+However, in many months during these years, the number of deaths remains notably high, often fluctuating between 10 and 15 deaths per month.
+Detailed discussion found in @sec-second-dpoint.
 ```{r}
@@
   ggplot(aes(x = YearMonth, y = total_decedents, group = 1)) +
-  geom_line(color = "blue") +
-  geom_point(color = "red") +
+  geom_line(color = "black") +
+  geom_point(color = "black") +
   scale_x_discrete(breaks = Death_data_filtered$YearMonth[seq(1, nrow(Death_data_filtered), by = 6)]) + # Show every 6th month
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) + # Display x-axis labels vertically
-  labs(title = "Trends in Total Decedents by Month and Year (2019-2023)",
+  labs(title = "Trends in Total Decedents by Month(2019-2023)",
        x = "Month-Year",
        y = "Total Decedents") +
-  theme_minimal()
+  theme_minimal()+
+  theme(plot.title = element_text(hjust = 0.5))
+
 ```
-
+\newpage
+@fig-T illustrates the number of deaths among genders.
+The bar graph clearly shows that the number of male decedents is significantly greater than that of female decedents, a trend consistent across all the years presented.
+Additionally, the graph indicates that the mortality rate among males is more volatile over the years, particularly from 2019 to 2022.
+These results may be attributed to the proportion of males living in shelters; further details can be found in @sec-3-dpoint.
 ```{r}
@@
   geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
-  labs(title = "Mortality Numbers of Male and Female Decedents",
+  labs(title = "Mortality Numbers of Male Vs Female Decedents over 2007 to 2024",
        x = "Year (2000s)",
@@
   scale_fill_manual(values = c("Male_decedents" = "blue", "Female_decedents" = "pink")) +
-  theme_minimal()
+  theme_minimal() +
+  theme(legend.title = element_blank(),
+        legend.position = "bottom") +
+  theme(plot.title = element_text(hjust = 0.5))
 
 ```
+@fig-AvgAgeofDecedentsFemaleVSMale further examines the differences in average age at death between females and males.
+It's important to note that there are significantly more male deaths than female deaths, which could result in a lower average age for males due to the larger sample size.
+Although the graph does not definitively indicate whether males generally die at an older age than females or vice versa, it does suggest that the average age at death for males appears to be considerably higher than that for females during many periods indicated in the graph.
+More information is available in @sec-3-dpoint.
 
@@
   geom_point(size = 2) +    
-  labs(title = "Average Age of Death for Male and Female Decedents",
+  labs(title = "Average Age of Death for Male Vs Female Decedents",
        x = "Year",
@@
   theme(legend.title = element_blank(),
-        legend.position = "bottom")
+        legend.position = "bottom") +
+  theme(plot.title = element_text(hjust = 0.5))
 
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 
-## First discussion point {#sec-first-point}
+## First discussion point {#sec-first-dpoint}
 @fig-TotalNumberofDeathByYear shows the total number of decedents among homeless individuals living in shelters.
@@
 The best way to prevent infection is to get vaccinated each year (@cdc_flu_key_facts).
-There is a clear trend of increasing deaths up until 2015, followed by a drop in 2016.
+Thus, the clear trend of increasing deaths from 2013 up until 2015, followed by a drop in 2016 shown in @fig-TotalNumberofDeathByYear perfectly aligns the duration of the outbreak.
 This pattern closely resembles the total number of influenza outbreaks reported in the article "Influenza Outbreaks in Ontario Hospitals" (@cameron2019).
-According to Table 1 in the article, the number of outbreaks increased from 38 in 2013-2014 to 117 in 2014-2015, then decreased to 36 in 2015-2016.
+According to Table 1 in the article, the number of outbreaks increased from 38 in '2013-2014' to 117 in '2014-2015', then decreased to 36 in '2015-2016'.
 This rise and fall in outbreaks corresponds with the trends observed in @fig-TotalNumberofDeathByYear, where the number of deaths peaked in 2015 and then significantly declined in 2016, coinciding with the flu being brought under control.
 
-One notable point is that in the article, the fatality rate for the 2014-2015 outbreak was only 2.4%, despite the hundreds of flu outbreaks that occurred that year.
+One notable point is that in the article (@cameron2019), the fatality rate for the 2014-2015 outbreak was only 2.4%, despite the hundreds of flu outbreaks that occurred that year.
 However,@fig-TotalNumberofDeathByYear shows that the number of deaths in 2015 was around 45.
@@
 
-## Second discussion point
+## Second discussion point {#sec-second-dpoint}
 
@@
 
-## Third discussion point
+## Third discussion point {#sec-3-dpoint}
 
@@
 
-# Appendix {-}
+# Appendix {#sec-app}
 
-# Additional data details
+```{r}
+#| label: fig-2007-2010
+#| fig-cap: Monthly Mortality rate from 2007-2010
+#| echo: false
+#| warning: false
+#| message: false
 
-# Model details {#sec-model-details}
+# Load necessary libraries
+library(ggplot2)
+library(dplyr)
+library(gridExtra)
 
-## Posterior predictive check
+Death_data$month <- factor(Death_data$month, levels = month.abb)
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
+# Filter data for the years 2007 and 2008
+current_year_data_2007 <- Death_data %>% filter(year == 2007)
+current_year_data_2008 <- Death_data %>% filter(year == 2008)
+current_year_data_2009 <- Death_data %>% filter(year == 2009)
+current_year_data_2010 <- Death_data %>% filter(year == 2010)
 
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+# Create the bar plots for 2007 and 2008
+plot_2007 <- ggplot(current_year_data_2007, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2007", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45))+
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
 
-## Diagnostics
+plot_2008 <- ggplot(current_year_data_2008, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2008", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45))+
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
 
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
+plot_2009 <- ggplot(current_year_data_2009, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2009", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
 
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
+plot_2010 <- ggplot(current_year_data_2010, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2010", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+# Arrange plots side by side
+grid.arrange(plot_2007, plot_2008, plot_2009, plot_2010, ncol = 2)
+
+```
+
+```{r}
+#| label: fig-2011-2014
+#| fig-cap: Monthly Mortality rate from 2011-2014
+#| echo: false
+#| warning: false
+#| message: false
+
+# Load necessary libraries
+library(ggplot2)
+library(dplyr)
+library(gridExtra)
+
+Death_data$month <- factor(Death_data$month, levels = month.abb)
+
+# Filter data for the years 
+current_year_data_2011 <- Death_data %>% filter(year == 2011)
+current_year_data_2012 <- Death_data %>% filter(year == 2012)
+current_year_data_2013 <- Death_data %>% filter(year == 2013)
+current_year_data_2014 <- Death_data %>% filter(year == 2014)
+
+plot_2011 <- ggplot(current_year_data_2011, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2011", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+plot_2012 <- ggplot(current_year_data_2012, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2012", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+plot_2013 <- ggplot(current_year_data_2013, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2013", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+plot_2014 <- ggplot(current_year_data_2014, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2014", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+# Arrange plots side by side
+grid.arrange(plot_2011, plot_2012, plot_2013, plot_2014, ncol = 2)
+
+```
+
+```{r}
+#| label: fig-2015-2018
+#| fig-cap: Monthly Mortality rate from 2015-2018
+#| echo: false
+#| warning: false
+#| message: false
+
+Death_data$month <- factor(Death_data$month, levels = month.abb)
+
+# Filter data for the years
+
+current_year_data_2015 <- Death_data %>% filter(year == 2015)
+current_year_data_2016 <- Death_data %>% filter(year == 2016)
+current_year_data_2017 <- Death_data %>% filter(year == 2017)
+current_year_data_2018 <- Death_data %>% filter(year == 2018)
+
+# Create the bar plots
+
+plot_2015 <- ggplot(current_year_data_2015, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2015", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+plot_2016 <- ggplot(current_year_data_2016, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2016", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+plot_2017 <- ggplot(current_year_data_2017, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2017", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+plot_2018 <- ggplot(current_year_data_2018, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2018", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+# Arrange plots side by side
+grid.arrange(plot_2015, plot_2016, plot_2017, plot_2018, ncol = 2)
+
+```
+
+```{r}
+#| label: fig-2019-2022
+#| fig-cap: Monthly Mortality rate from 2019-2022
+#| echo: false
+#| warning: false
+#| message: false
+
+Death_data$month <- factor(Death_data$month, levels = month.abb)
+
+# Filter data for the years
+
+current_year_data_2019 <- Death_data %>% filter(year == 2019)
+current_year_data_2020 <- Death_data %>% filter(year == 2020)
+current_year_data_2021 <- Death_data %>% filter(year == 2021)
+current_year_data_2022 <- Death_data %>% filter(year == 2022)
+
+# Create the bar plots
+plot_2019 <- ggplot(current_year_data_2019, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2019", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+plot_2020 <- ggplot(current_year_data_2020, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2020", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+plot_2021 <- ggplot(current_year_data_2021, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2021", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+plot_2022 <- ggplot(current_year_data_2022, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2022", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+# Arrange plots side by side
+grid.arrange(plot_2019, plot_2020, plot_2021, plot_2022, ncol = 2)
+
+```
+
+```{r}
+#| label: fig-2023-2024
+#| fig-cap: Monthly Mortality rate from 2023-2024
+#| echo: false
+#| warning: false
+#| message: false
+
+Death_data$month <- factor(Death_data$month, levels = month.abb)
+
+# Filter data for the years 
+current_year_data_2023 <- Death_data %>% filter(year == 2023)
+current_year_data_2024 <- Death_data %>% filter(year == 2024)
+
+# Create the bar plots
+plot_2023 <- ggplot(current_year_data_2023, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2023", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+plot_2024 <- ggplot(current_year_data_2024, aes(x = month, y = total_decedents)) +
+  geom_bar(stat = "identity", fill = "black") +
+  labs(title = "Total Decedents by Month in 2024", 
+       x = "Month", 
+       y = "# Decedents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45)) +
+  theme(plot.title = element_text(hjust = 0.5, size = 10))
+
+# Arrange plots side by side
+grid.arrange(plot_2023, plot_2024, ncol = 2)
+
+```
 
 \newpage
-
 # References

### 2024-09-27 scripts/01-download_data.R
delete unnecessary comment line

@@ -35,8 +35,6 @@ Avg_death_age_data <- filter(datastore_resources, row_number()==2) %>% get_resou
 
 
 #### Save data ####
-# [...UPDATE THIS...]
-# change the_raw_data to whatever name you assigned when you downloaded it.
 write_csv(Num_death_data, "data/raw_data/Num_death_raw_data.csv") 
 write_csv(Avg_death_age_data, "data/raw_data/raw_age_data.csv") 
 

### 2024-09-27 paper/paper.qmd
Updated and cite R function used

@@
 library(dplyr)
+library(knitr)
+library(tidyr)
+library(gridExtra)
 
@@
 The data used in this paper is access in from Open Data Toronto and the particular data set used was the Deaths of Shelter Residents (@toronto_deaths_shelter_residents).
-To analysis the data and creating graphs using the data, following package that was build in the (R program @citeR) was used: tidyverse (@tidyverse), dplyr (@dplyr), knitr (@knitr), and ggplot2 (@ggplot2).
+To analysis the data and creating graphs using the data, following package that was build in the (R program @citeR) was used: tidyverse (@tidyverse), dplyr (@dplyr), tidyr (@tidyr), gridExtra (@gridExtra), knitr (@knitr), and ggplot2 (@ggplot2).
 LLM (@openai2023chatgpt) was also use for this paper in term of helping with coding in R to graph data, simulating a data, cleaning data, and revising writing.
@@
 #| echo: false
-
-library(knitr)
-library(dplyr)
 
@@
 #| message: false
-
-library(ggplot2)
-library(tidyr)
-library(dplyr)
 
@@
 
-# Load necessary libraries
-library(ggplot2)
-library(dplyr)
-library(gridExtra)
-
 Death_data$month <- factor(Death_data$month, levels = month.abb)
@@
 #| message: false
-
-# Load necessary libraries
-library(ggplot2)
-library(dplyr)
-library(gridExtra)
 

### 2024-09-27 scripts/00-simulate_data.R
edit the header info

@@ -1,11 +1,10 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
+# Purpose: Simulates a sample data of Deaths Among Shelter Residents from 2007 to 2024
 # Author: Mandy He
 # Date: 23 September 2024 
 # Contact: mandyxy.he@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: 
-# Any other information needed? 
+# License: N/A
+# Pre-requisites: N/A
 
 
 #### Workspace setup ####

### 2024-09-27 scripts/01-download_data.R

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
+# Purpose: Downloads and saves the data from Open Data Toronto
 # Author: Mandy He
 # Date: 23 September 2024 
 # Contact: mandyxy.he@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# License: N/A
+# Pre-requisites: N/A
+
 
 
 #### Workspace setup ####

### 2024-09-27 scripts/02-data_cleaning.R

@@ -1,11 +1,10 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Cleans the raw data so variable is easier to call 
+# Author: Mandy He
+# Date: 23 September 2024 
+# Contact: mandyxy.he@mail.utoronto.ca
+# License: N/A
+# Pre-requisites: N/A
 
 #### Workspace setup ####
 library(tidyverse)

### 2024-09-27 scripts/03-test_data.R

@@ -1,11 +1,10 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Tests for "n/a" Replacement and Negative Numbers.
+# Author: Mandy He
+# Date: 23 September 2024 
+# Contact: mandyxy.he@mail.utoronto.ca
+# License: N/A
+# Pre-requisites: N/A
 
 
 #### Workspace setup ####

### 2024-09-27 paper/paper.qmd
change title and intro to better fit the rubric

@@
 ---
-title: "Trends in Mortality Rates"
-subtitle: "Analyzing Deaths Among Shelter Residents from 2007 to 2024"
+title: "Trends in Mortality Among Shelter Residents: : Closely Tied To Virus Outbreaks"
+subtitle: "Analyzing period: 2007 to 2024"
 author: 
@@
 thanks: "Code and data are available at: https://open.toronto.ca/dataset/deaths-of-shelter-residents/"
-date: Septmeber 23 2024
+date: Septmeber 27 2024
 date-format: long
-abstract: "Poverty and homelessness remain pressing issues in today’s society, particularly evident in downtown Toronto, where many individuals still sleep on the streets or in subway stations.
-Due to these harsh living conditions, these individuals often face severe health issues, leading to an increased risk of mortality.
-Luckily, there are services in the city that provide place for some of these people to live.
-This paper analyzes the trends in mortality rates among shelter residents in the City of Toronto from 2007 to 2024, highlighting patterns over time and exploring the factors contributing to these rates."
+abstract: "Homelessness remain pressing issues in today’s society, particularly evident in downtown Toronto, due to  harsh living conditions, these individuals often face severe health issues, leading to an increased risk of mortality.
+This paper analyzes the trends in mortality rates among shelter residents in the City of Toronto from 2007 to 2024, highlighting patterns over time and exploring the factors contributing to these rates.
+This paper found that the trend in the mortality rate of this minority group reflects the overall health conditions in the macro-environment.
+Therefore, examining these patterns can help improve provisions and reduce the number of deaths among both this vulnerable population and society at large."
 format: pdf
@@
 
+#load necessary package 
 library(tidyverse)
@@
 
+# load data
 Death_data <- read_csv(here::here("data/analysis_data/Num_death_analysis_data.csv"))
@@
 Additionally, the average age at death is around 50 to 60 years old, which is relatively young compared to Canada’s life expectancy.
-Future details about the findings will be discussed in the discussion section.
+These trends are crucial for helping the city improve services provided to this population.
+Furthermore, from a social and economic perspective, supporting minority populations is essential for community and economic growth.
+If individuals in these communities are healthier and live longer, they can contribute more effectively to society.
 
@@
 Shelters are required to notify the City of Toronto within 24 hours of a death and submit a written report within 30 days.
-Consequently, the data is updated monthly on the Open Data Toronto website.
+Accordingly, the data is updated monthly on the Open Data Toronto website.
 The raw data set includes 10 variables and 18 observations, covering the period from 2007 to 2024, as shown in @tbl-Table1.
@@
 #| label: tbl-Table1
-#| tbl-cap: Trends in Mortality Rates Among Shelter Residents (2007-2024)
+#| tbl-cap: Variable In The Data - Detah Among Shelter Residents (2007-2024)
 #| echo: false
@@
 
+# creating a line graph 
 AvgAge_data |> 
@@
 #| label: fig-TotalDecedentbyMonthcovid
-#| fig-cap: Yearly Mortality rate from 2019-2023
+#| fig-cap: Yearly Mortality rate during COVID (2019-2023)
 #| echo: false
@@
 # Filter the data for years 2019 to 2022
-Death_data_filtered <- Death_data %>%
+Death_data_filtered <- Death_data |> 
   filter(year >= 2019 & year <= 2023)
@@
   geom_point(color = "black") +
-  scale_x_discrete(breaks = Death_data_filtered$YearMonth[seq(1, nrow(Death_data_filtered), by = 6)]) + # Show every 6th month
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) + # Display x-axis labels vertically
+  scale_x_discrete(breaks = Death_data_filtered$YearMonth[seq(1, nrow(Death_data_filtered), by = 6)]) + 
   labs(title = "Trends in Total Decedents by Month(2019-2023)",
@@
   theme_minimal()+
+  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) + 
   theme(plot.title = element_text(hjust = 0.5))
@@
 #| label: fig-T 
-#| fig-cap: Total Decedents Female VS Male
+#| fig-cap: Total Decedents Split by Gender (Female VS Male)
 #| echo: false
@@
 # Reshape the data for plotting
-plot_data <- AvgAge_data %>%
-  select(year, Male_decedents, Female_decedents) %>%
+plot_data <- AvgAge_data |>
+  select(year, Male_decedents, Female_decedents)|>
   pivot_longer(cols = c(Male_decedents, Female_decedents),
@@
 # Create a new variable for the last two digits of the year with leading zero
-plot_data <- plot_data %>%
+plot_data <- plot_data |>
   mutate(year_short = sprintf("%02d", year %% 100))
@@
 # Create the bar graph with adjusted year labels
-ggplot(plot_data, aes(x = year_short, y = Mortality, fill = Gender)) +
+plot_data |>
+  ggplot(aes(x = year_short, y = Mortality, fill = Gender)) +
   geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
@@
 #| label: fig-AvgAgeofDecedentsFemaleVSMale 
-#| fig-cap: Total Decedents Female VS Male
+#| fig-cap: Average Age of Death (Female VS Male)
 #| echo: false
@@
 # Reshape the data for plotting average age of death
-avg_age_data <- AvgAge_data %>%
+avg_age_data <- AvgAge_data |>
   select(year, Average_age_of_death_male_decedents, Average_age_of_death_female_decedents) %>%
@@
                names_to = "Gender",
-               values_to = "Average_Age") %>%
+               values_to = "Average_Age") |>
   mutate(year_short = sprintf("%02d", year %% 100))
@@
 # Create the line graph for average age of death
-ggplot(avg_age_data, aes(x = year_short, y = Average_Age, color = Gender, group = Gender)) +
+avg_age_data |>
+  ggplot(aes(x = year_short, y = Average_Age, color = Gender, group = Gender)) +
   geom_line(size = 1) +   
@@
 The findings in this paper represent a small aspect of what is happening in the world, focusing specifically on the homeless population.
-While these results cannot be used to draw definitive conclusions about broader trends, they do provide insights into patterns that may be occurring on a larger scale.
+While these results cannot be used to draw definitive conclusions about broader trends, they do provide awareness into patterns that may be occurring on a larger scale.
 
@@
 
+# make sure that the month is in order from Jan to Dec.
 Death_data$month <- factor(Death_data$month, levels = month.abb)
@@
 
+# Create the bar plots
 plot_2011 <- ggplot(current_year_data_2011, aes(x = month, y = total_decedents)) +

### 2024-09-27 paper/paper.qmd
Made change to the coding style using styler

@@
 ---
-title: "Trends in Mortality Among Shelter Residents: : Closely Tied To Virus Outbreaks"
-subtitle: "Analyzing period: 2007 to 2024"
+title: "Mortality Trend Among Shelter Residents: Closely Tied To Virus Outbreaks"
 author: 
@@
 date-format: long
-abstract: "Homelessness remain pressing issues in today’s society, particularly evident in downtown Toronto, due to  harsh living conditions, these individuals often face severe health issues, leading to an increased risk of mortality.
-This paper analyzes the trends in mortality rates among shelter residents in the City of Toronto from 2007 to 2024, highlighting patterns over time and exploring the factors contributing to these rates.
-This paper found that the trend in the mortality rate of this minority group reflects the overall health conditions in the macro-environment.
+abstract: "Homelessness remain pressing issues in today’s society, particularly evident in downtown Toronto, these individuals often face severe health issues, leading to an increased risk of mortality.
+This paper analyzes trends in mortality rates among shelter residents in the City of Toronto from 2007 to 2024, highlighting patterns over time and exploring the factors contributing to these rates.
+This paper found trend in mortality rate of this minority group reflects the overall health conditions in the macro-environment.
 Therefore, examining these patterns can help improve provisions and reduce the number of deaths among both this vulnerable population and society at large."
@@
 
-#load necessary package 
+# load necessary package
 library(tidyverse)
@@
 AvgAge_data <- read_csv(here::here("data/analysis_data/AvgAge_death_analysis_data.csv"))
-
 ```
@@
 kable(AvgAge_data_no_id,
-      col.names = c("Year", "Total Decedents", 
+  col.names = c(
+    "Year", "Total Decedents",
     "Avg Age of Death (All)", "Male Decedents",
@@
     "Avg Age of Death (Female)", "Trans/NB/2S Decedents",
-                    "Avg Age of Death (Trans/NB/2S)"))
+    "Avg Age of Death (Trans/NB/2S)"
+  )
+)
 ```
@@
   scale_y_continuous(breaks = seq(0, max(AvgAge_data$total_decedents), by = 20))
-
 ```
@@
   scale_x_discrete(breaks = Death_data_filtered$YearMonth[seq(1, nrow(Death_data_filtered), by = 6)]) +
-  labs(title = "Trends in Total Decedents by Month(2019-2023)",
+  labs(
+    title = "Trends in Total Decedents by Month(2019-2023)",
     x = "Month-Year",
-       y = "Total Decedents") +
-  theme_minimal()+
+    y = "Total Decedents"
+  ) +
+  theme_minimal() +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) +
   theme(plot.title = element_text(hjust = 0.5))
-
 ```
@@
 plot_data <- AvgAge_data |>
-  select(year, Male_decedents, Female_decedents)|>
-  pivot_longer(cols = c(Male_decedents, Female_decedents),
+  select(year, Male_decedents, Female_decedents) |>
+  pivot_longer(
+    cols = c(Male_decedents, Female_decedents),
     names_to = "Gender",
-               values_to = "Mortality")
+    values_to = "Mortality"
+  )
 
@@
   geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
-  labs(title = "Mortality Numbers of Male Vs Female Decedents over 2007 to 2024",
+  labs(
+    title = "Mortality Numbers of Male Vs Female Decedents over 2007 to 2024",
     x = "Year (2000s)",
-       y = "Number of Decedents") +
+    y = "Number of Decedents"
+  ) +
   scale_fill_manual(values = c("Male_decedents" = "blue", "Female_decedents" = "pink")) +
   theme_minimal() +
-  theme(legend.title = element_blank(),
-        legend.position = "bottom") +
+  theme(
+    legend.title = element_blank(),
+    legend.position = "bottom"
+  ) +
   theme(plot.title = element_text(hjust = 0.5))
-
 ```
@@
   select(year, Average_age_of_death_male_decedents, Average_age_of_death_female_decedents) %>%
-  pivot_longer(cols = c(Average_age_of_death_male_decedents, Average_age_of_death_female_decedents),
+  pivot_longer(
+    cols = c(Average_age_of_death_male_decedents, Average_age_of_death_female_decedents),
     names_to = "Gender",
-               values_to = "Average_Age") |>
+    values_to = "Average_Age"
+  ) |>
   mutate(year_short = sprintf("%02d", year %% 100))
@@
   geom_point(size = 2) +
-  labs(title = "Average Age of Death for Male Vs Female Decedents",
+  labs(
+    title = "Average Age of Death for Male Vs Female Decedents",
     x = "Year",
-       y = "Average Age of Death") +
-  scale_color_manual(values = c("Average_age_of_death_male_decedents" = "blue", 
-                                 "Average_age_of_death_female_decedents" = "pink")) +
+    y = "Average Age of Death"
+  ) +
+  scale_color_manual(values = c(
+    "Average_age_of_death_male_decedents" = "blue",
+    "Average_age_of_death_female_decedents" = "pink"
+  )) +
   theme_minimal() +
-  theme(legend.title = element_blank(),
-        legend.position = "bottom") +
+  theme(
+    legend.title = element_blank(),
+    legend.position = "bottom"
+  ) +
   theme(plot.title = element_text(hjust = 0.5))
-
 ```
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2007", 
+  labs(
+    title = "Total Decedents by Month in 2007",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45))+
+  theme(axis.text.x = element_text(angle = 45)) +
   theme(plot.title = element_text(hjust = 0.5, size = 10))
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2008", 
+  labs(
+    title = "Total Decedents by Month in 2008",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45))+
+  theme(axis.text.x = element_text(angle = 45)) +
   theme(plot.title = element_text(hjust = 0.5, size = 10))
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2009", 
+  labs(
+    title = "Total Decedents by Month in 2009",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2010", 
+  labs(
+    title = "Total Decedents by Month in 2010",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
 grid.arrange(plot_2007, plot_2008, plot_2009, plot_2010, ncol = 2)
-
 ```
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2011", 
+  labs(
+    title = "Total Decedents by Month in 2011",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2012", 
+  labs(
+    title = "Total Decedents by Month in 2012",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2013", 
+  labs(
+    title = "Total Decedents by Month in 2013",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2014", 
+  labs(
+    title = "Total Decedents by Month in 2014",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
 grid.arrange(plot_2011, plot_2012, plot_2013, plot_2014, ncol = 2)
-
 ```
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2015", 
+  labs(
+    title = "Total Decedents by Month in 2015",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2016", 
+  labs(
+    title = "Total Decedents by Month in 2016",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2017", 
+  labs(
+    title = "Total Decedents by Month in 2017",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2018", 
+  labs(
+    title = "Total Decedents by Month in 2018",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
 grid.arrange(plot_2015, plot_2016, plot_2017, plot_2018, ncol = 2)
-
 ```
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2019", 
+  labs(
+    title = "Total Decedents by Month in 2019",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2020", 
+  labs(
+    title = "Total Decedents by Month in 2020",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2021", 
+  labs(
+    title = "Total Decedents by Month in 2021",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2022", 
+  labs(
+    title = "Total Decedents by Month in 2022",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
 grid.arrange(plot_2019, plot_2020, plot_2021, plot_2022, ncol = 2)
-
 ```
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2023", 
+  labs(
+    title = "Total Decedents by Month in 2023",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
   geom_bar(stat = "identity", fill = "black") +
-  labs(title = "Total Decedents by Month in 2024", 
+  labs(
+    title = "Total Decedents by Month in 2024",
     x = "Month",
-       y = "# Decedents") +
+    y = "# Decedents"
+  ) +
   theme_minimal() +
@@
 grid.arrange(plot_2023, plot_2024, ncol = 2)
-
 ```

### 2024-09-27 scripts/00-simulate_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Simulates a sample data of Deaths Among Shelter Residents from 2007 to 2024
 # Author: Mandy He
-# Date: 23 September 2024 
+# Date: 23 September 2024
 # Contact: mandyxy.he@mail.utoronto.ca
 # License: N/A
 # Pre-requisites: N/A
@@ -18,7 +18,7 @@ set.seed(304)
 # Set parameters for the simulation
 start_year <- 2007
 end_year <- 2024
-months <- month.abb  # Built-in R vector for month abbreviations
+months <- month.abb # Built-in R vector for month abbreviations
 
 # Initialize an empty data frame to hold the data
 data <- data.frame()
@@ -27,20 +27,20 @@ data <- data.frame()
 for (year in start_year:end_year) {
   for (month in months) {
     # Randomly determine the total number of decedents for the month
-    total_decedents <- rpois(1, lambda = 2)  # Using Poisson distribution for count data
-    
+    total_decedents <- rpois(1, lambda = 2) # Using Poisson distribution for count data
+
     # Ensure total decedents is not negative
     total_decedents <- max(total_decedents, 0)
-    
+
     # Randomly assign gender categories based on proportions
     if (total_decedents > 0) {
-      male <- sample(0:total_decedents, 1)  # Randomly assign number of males
-      female <- total_decedents - male        # The rest will be females
-      transgender_non_binary <- sample(0:1, 1)  # Randomly assign Transgender/Non-binary/Two-Spirit
+      male <- sample(0:total_decedents, 1) # Randomly assign number of males
+      female <- total_decedents - male # The rest will be females
+      transgender_non_binary <- sample(0:1, 1) # Randomly assign Transgender/Non-binary/Two-Spirit
     } else {
       male <- female <- transgender_non_binary <- 0
     }
-    
+
     # Create a new row of data
     new_row <- data.frame(
       id = nrow(data) + 1,
@@ -51,7 +51,7 @@ for (year in start_year:end_year) {
       Female = female,
       Transgender_Non_binary_Two_Spirit = transgender_non_binary
     )
-    
+
     # Append the new row to the data frame
     data <- bind_rows(data, new_row)
   }
@@ -66,5 +66,3 @@ dir.create("data/raw_data", recursive = TRUE)
 
 # Write the data to CSV
 write_csv(data, file = "data/raw_data/simulated.csv")
-
-

### 2024-09-27 scripts/01-download_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Downloads and saves the data from Open Data Toronto
 # Author: Mandy He
-# Date: 23 September 2024 
+# Date: 23 September 2024
 # Contact: mandyxy.he@mail.utoronto.ca
 # License: N/A
 # Pre-requisites: N/A
@@ -24,18 +24,16 @@ package
 resources <- list_package_resources("deaths-of-shelter-residents")
 
 # identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
+datastore_resources <- filter(resources, tolower(format) %in% c("csv", "geojson"))
 
 # load the Deaths of Shelter Residents data
-Num_death_data <- filter(datastore_resources, row_number()==1) %>% get_resource()
+Num_death_data <- filter(datastore_resources, row_number() == 1) %>% get_resource()
 
 
 # Deaths of Shelter Residents Mean Age
-Avg_death_age_data <- filter(datastore_resources, row_number()==2) %>% get_resource()
+Avg_death_age_data <- filter(datastore_resources, row_number() == 2) %>% get_resource()
 
 
 #### Save data ####
-write_csv(Num_death_data, "data/raw_data/Num_death_raw_data.csv") 
-write_csv(Avg_death_age_data, "data/raw_data/raw_age_data.csv") 
-
-         
+write_csv(Num_death_data, "data/raw_data/Num_death_raw_data.csv")
+write_csv(Avg_death_age_data, "data/raw_data/raw_age_data.csv")

### 2024-09-27 scripts/02-data_cleaning.R

@@ -1,7 +1,7 @@
 #### Preamble ####
-# Purpose: Cleans the raw data so variable is easier to call 
+# Purpose: Cleans the raw data so variable is easier to call
 # Author: Mandy He
-# Date: 23 September 2024 
+# Date: 23 September 2024
 # Contact: mandyxy.he@mail.utoronto.ca
 # License: N/A
 # Pre-requisites: N/A
@@ -15,7 +15,7 @@ library(readr)
 
 #### Clean data ####
 
-## clean the number of death data frist 
+## clean the number of death data frist
 raw_death_data <- read_csv("data/raw_data/Num_death_raw_data.csv")
 
 # Display the first few rows of the original data set
@@ -33,10 +33,8 @@ cleaned_death_data <- raw_death_data %>%
     female = `Female`,
     transgender_nonbinary_twospirit = `Transgender/Non-binary/Two-Spirit`
   ) %>%
-  
   # Replace "n/a" with 0 in the relevant columns
   mutate(across(everything(), ~ ifelse(. == "n/a", 0, .))) %>%
-  
   # Convert relevant columns to numeric
   mutate(
     id = as.integer(id),
@@ -51,14 +49,7 @@ cleaned_death_data <- raw_death_data %>%
 # Display the cleaned dataset
 head(cleaned_death_data)
 
-#Key Changes: Backticks: I added backticks around all column names that contain spaces or special characters (e.g., Total decedents, Transgender/Non-binary/Two-Spirit).
-#Usage:Replace the File Path: Ensure to replace "path_to_your_file.csv" with the path to your actual dataset.
-#Run the Script: Execute the script, and it should run without errors.
-
-
-
-
-## Clean the Average age of death data. 
+## Clean the Average age of death data.
 raw_age_data <- read_csv("data/raw_data/raw_age_data.csv")
 
 # Display the first few rows of the original dataset
@@ -66,7 +57,6 @@ head(raw_age_data)
 
 # Clean the data
 cleaned_avg_age_death_data <- raw_age_data %>%
-  
   # Rename columns to remove spaces and special characters
   rename(
     id = `_id`,
@@ -80,14 +70,12 @@ cleaned_avg_age_death_data <- raw_age_data %>%
     Transgender_Non_binary_Two_Spirit_decedents = `Transgender/Non-binary/Two-Spirit decedents`,
[…]

### 2024-09-27 scripts/03-test_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Tests for "n/a" Replacement and Negative Numbers.
 # Author: Mandy He
-# Date: 23 September 2024 
+# Date: 23 September 2024
 # Contact: mandyxy.he@mail.utoronto.ca
 # License: N/A
 # Pre-requisites: N/A
@@ -17,7 +17,7 @@ library(tidyverse)
 test_na_replacement <- function(cleaned_avg_age_death_data) {
   # Check if any "n/a" values remain in the data
   has_na <- any(cleaned_avg_age_death_data == "n/a")
-  
+
   if (!has_na) {
     print("Test 1 Passed: All 'n/a' values have been replaced with 0.")
   } else {
@@ -43,10 +43,10 @@ test_negative_numbers <- function(cleaned_avg_age_death_data) {
     "Transgender/Non-binary/Two-Spirit decedents",
     "Avg age of death, transgender/non-binary/two-spirit decedents"
   )
-  
+
   # Check for negative values in the specified columns
   has_negative <- any(sapply(cleaned_avg_age_death_data[numeric_columns], function(x) any(x < 0)))
-  
+
   if (!has_negative) {
     print("Negative Number Test Passed: No negative values found in the relevant columns.")
   } else {
@@ -67,10 +67,10 @@ test_negative_numbers <- function(cleaned_death_data) {
     "Female",
     "Transgender/Non-binary/Two-Spirit"
   )
-  
+
   # Check for negative values in the specified columns
   has_negative <- any(sapply(cleaned_death_data[numeric_columns], function(x) any(x < 0)))
-  
+
   if (!has_negative) {
     print("Negative Number Test Passed: No negative values found in the relevant columns.")
   } else {

### 2024-09-27 scripts/04-Code_auto_format.R

@@ -1,0 +1,9 @@
+# Load the styler package
+library(styler)
+
+# Define files to format
+files_to_format <- c("scripts/00-simulate_data.R", "scripts/01-download_data.R", "scripts/02-data_cleaning.R",
+                     "scripts/03-test_data.R", "paper/paper.qmd")
+
+# Loop through each file and format it
+lapply(files_to_format, style_file)

### 2024-09-27 paper/paper.qmd
Change the format of the paper for readability

@@
 ---
-title: "Mortality Trend Among Shelter Residents: Closely Tied To Virus Outbreaks"
+title: "Mortality Trend Among Shelter Residents: "
+subtitle: "Closely Tied To Virus Outbreaks"
 author: 
@@
 date-format: long
-abstract: "Homelessness remain pressing issues in today’s society, particularly evident in downtown Toronto, these individuals often face severe health issues, leading to an increased risk of mortality.
-This paper analyzes trends in mortality rates among shelter residents in the City of Toronto from 2007 to 2024, highlighting patterns over time and exploring the factors contributing to these rates.
-This paper found trend in mortality rate of this minority group reflects the overall health conditions in the macro-environment.
-Therefore, examining these patterns can help improve provisions and reduce the number of deaths among both this vulnerable population and society at large."
+abstract: "Homelessness remain pressing issues in today’s society, particularly evident in Toronto, these individuals face severe health issues, leading to increased risk of mortality.
+This paper analyzes trends in mortality rates among shelter residents in the City of Toronto (2007 to 2024), highlighting patterns over time and exploring the factors contributing to these rates.
+This paper found trend in mortality rate of this minority group reflects the overall macro-environment health conditions.
+Thus, examining these patterns can help improve provisions and reduce the number of deaths among both this vulnerable population and society at large."
 format: pdf
@@
 One notable point is that in the article (@cameron2019), the fatality rate for the 2014-2015 outbreak was only 2.4%, despite the hundreds of flu outbreaks that occurred that year.
-However,@fig-TotalNumberofDeathByYear shows that the number of deaths in 2015 was around 45.
+However, @fig-TotalNumberofDeathByYear shows that the number of deaths in 2015 was around 45.
 This discrepancy can be explained by the fact that the data in the article pertains to hospital outbreaks, where individuals have access to better medical care.

### 2024-09-27 paper/paper.qmd
Link to github in footnote

@@
   - Mandy He
-thanks: "Code and data are available at: https://open.toronto.ca/dataset/deaths-of-shelter-residents/"
+thanks: "Code and data are available at: https://github.com/MandyHe7/Mortality-Trend-Among-Shelter-Residents"
 date: Septmeber 27 2024

### 2024-09-27 paper/paper.qmd
final preread of the paper and edits

@@
 
-In this paper we will look at the data @toronto_deaths_shelter_residents, which explore the trends in mortality rates among shelter residents in the City of Toronto from 2007 to 2024, highlighting patterns over time and exploring the factors contributing to these rates.The data indicates that mortality rates were relatively low, with fewer than five deaths per month in the prior 2019.
+In this paper we will look at a data (@toronto_deaths_shelter_residents), which explore the trends in mortality rates among shelter residents in the City of Toronto from 2007 to 2024, highlighting patterns over time and exploring the factors contributing to these rates.
+The data indicates that mortality rates were relatively low, with fewer than five deaths per month prior of 2019.
 However, after 2019, there was a sharp increase in the number of deaths, peaking in 2021 and 2022.
@@
 
-The data used in this paper is access in from Open Data Toronto and the particular data set used was the Deaths of Shelter Residents (@toronto_deaths_shelter_residents).
+The data used in this paper is access from Open Data Toronto and the particular data set used was the Deaths of Shelter Residents (@toronto_deaths_shelter_residents).
 To analysis the data and creating graphs using the data, following package that was build in the (R program @citeR) was used: tidyverse (@tidyverse), dplyr (@dplyr), tidyr (@tidyr), gridExtra (@gridExtra), knitr (@knitr), and ggplot2 (@ggplot2).
-LLM (@openai2023chatgpt) was also use for this paper in term of helping with coding in R to graph data, simulating a data, cleaning data, and revising writing.
+LLM (@openai2023chatgpt) was also use in term of helping with coding in R to graph data, simulating data, cleaning data, and revising writing.
 
@@
 
-The Deaths of Shelter Residents data is part of a larger database shared by Toronto Public Health (TPH), which also includes deaths of homeless individuals living in shelters and outdoors (@toronto_homeless_deaths).
-Therefore, the findings in this paper focus solely on homeless individuals residing in shelters, which may limit the data set and lead to potential biases that do not fully represent the mortality of all homeless individuals.
+The Deaths of Shelter Residents data is part of a larger database shared by Toronto Public Health (TPH), which includes deaths of homeless individuals living in shelters and outdoors (@toronto_homeless_deaths).
+Therefore, the findings in this paper focus solely on homeless individuals residing in shelters, which limit the data set and lead to potential biases that do not fully represent the mortality of all homeless individuals.
 
@@
 The following graph (@fig-TotalNumberofDeathByYear) shows the overall trend in the number of decedents living in shelters from 2007 to 2024.
-It reveals that between 2007 and 2013, the number of deaths remained quite low, under 30 each year.
+It reveals that between 2007 and 2013, the number of deaths remained quite low, under 30 death each year.
 However, starting in 2013, the number of deaths increased, and the mortality rate never returned to the low levels seen from 2007 to 2013.
@@
 @fig-AvgAgeofDecedentsFemaleVSMale further examines the differences in average age at death between females and males.
-It's important to note that there are significantly more male deaths than female deaths, which could result in a lower average age for males due to the larger sample size.
+It is important to note that there are significantly more male deaths than female deaths, which could result in a lower average age for males due to the larger sample size.
 Although the graph does not definitively indicate whether males generally die at an older age than females or vice versa, it does suggest that the average age at death for males appears to be considerably higher than that for females during many periods indicated in the graph.
@@
 A model should be run to test the relationship between these variables.
+
+In summary, the paper does not provide statistical evidence for all the possible reasons discussed regarding the mortality trends of homeless individuals residing in shelters.
+However, it can serve as a guide and signal for city organizations to enhance their efforts in supporting the homeless population.
 

### 2024-09-27 paper/paper.qmd
made change to graph title font size

@@
   ) +
-  theme(plot.title = element_text(hjust = 0.5))
+  theme(plot.title = element_text(hjust = 0.5,size = 10))
 ```
@@
 #| label: fig-AvgAgeofDecedentsFemaleVSMale
-#| fig-cap: Average Age of Death (Female VS Male)
+#| fig-cap: Average Age of Death (Female VS Male) 2007-2024
 #| echo: false
@@
     title = "Average Age of Death for Male Vs Female Decedents",
-    x = "Year",
+    x = "Year (in 2000s)",
     y = "Average Age of Death"

### 2025-06-20 scripts/00-simulate_data.R
scripts from R to python

@@ -1,68 +1,0 @@
-#### Preamble ####
-# Purpose: Simulates a sample data of Deaths Among Shelter Residents from 2007 to 2024
-# Author: Mandy He
-# Date: 23 September 2024
-# Contact: mandyxy.he@mail.utoronto.ca
-# License: N/A
-# Pre-requisites: N/A
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(dplyr)
-
-
-#### Simulate data ####
-set.seed(304)
-
-# Set parameters for the simulation
-start_year <- 2007
-end_year <- 2024
-months <- month.abb # Built-in R vector for month abbreviations
-
-# Initialize an empty data frame to hold the data
-data <- data.frame()
-
-# Loop through the years and months to generate data
-for (year in start_year:end_year) {
-  for (month in months) {
-    # Randomly determine the total number of decedents for the month
-    total_decedents <- rpois(1, lambda = 2) # Using Poisson distribution for count data
-
-    # Ensure total decedents is not negative
-    total_decedents <- max(total_decedents, 0)
-
-    # Randomly assign gender categories based on proportions
-    if (total_decedents > 0) {
-      male <- sample(0:total_decedents, 1) # Randomly assign number of males
-      female <- total_decedents - male # The rest will be females
-      transgender_non_binary <- sample(0:1, 1) # Randomly assign Transgender/Non-binary/Two-Spirit
-    } else {
-      male <- female <- transgender_non_binary <- 0
-    }
-
-    # Create a new row of data
-    new_row <- data.frame(
-      id = nrow(data) + 1,
-      Year = year,
-      Month = month,
-      Total_decedents = total_decedents,
-      Male = male,
-      Female = female,
-      Transgender_Non_binary_Two_Spirit = transgender_non_binary
-    )
-
-    # Append the new row to the data frame
-    data <- bind_rows(data, new_row)
-  }
-}
-
-# Display the first few rows of the simulated data
-head(data)
-
-
-# Create the directory if it doesn't exist
-dir.create("data/raw_data", recursive = TRUE)
-
-# Write the data to CSV
-write_csv(data, file = "data/raw_data/simulated.csv")

### 2025-06-20 scripts/00-simulate_data.py

@@ -1,0 +1,63 @@
+#### Preamble ####
+# Purpose: Simulates a sample data of Deaths Among Shelter Residents from 2007 to 2024
+# Author: Mandy He
+# Date: 23 September 2024
+# Contact: mandyxy.he@mail.utoronto.ca
+# License: N/A
+# Pre-requisites: N/A
+
+#### Workspace setup ####
+import pandas as pd
+import numpy as np
+import os
+
+#### Simulate data ####
+np.random.seed(304)
+
+# Set parameters for the simulation
+start_year = 2007
+end_year = 2024
+months = ["Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
+
+# Initialize an empty list to hold the data
+data = []
+
+# Loop through the years and months to generate data
+for year in range(start_year, end_year + 1):
+    for month in months:
+        # Randomly determine the total number of decedents for the month
+        total_decedents = np.random.poisson(lam=2)
+
+        # Ensure total decedents is not negative
+        total_decedents = max(total_decedents, 0)
+
+        # Randomly assign gender categories based on proportions
+        if total_decedents > 0:
+            male = np.random.randint(0, total_decedents + 1)  # Random number of males
+            female = total_decedents - male  # Remainder are females
+            transgender_non_binary = np.random.randint(0, 2)  # 0 or 1
+        else:
+            male = female = transgender_non_binary = 0
+
+        # Append the new row to the data list
+        data.append({
+            "id": len(data) + 1,
+            "Year": year,
+            "Month": month,
+            "Total_decedents": total_decedents,
+            "Male": male,
+            "Female": female,
+            "Transgender_Non_binary_Two_Spirit": transgender_non_binary
+        })
+
+# Convert to DataFrame
+df = pd.DataFrame(data)
+
+# Display the first few rows of the simulated data
+print(df.head())
+
+# Create the directory if it doesn't exist
+os.makedirs("data/raw_data", exist_ok=True)
+
+# Write the data to CSV
+df.to_csv("data/raw_data/simulated.csv", index=False)

### 2025-06-20 scripts/01-download_data.R

@@ -1,39 +1,0 @@
-#### Preamble ####
-# Purpose: Downloads and saves the data from Open Data Toronto
-# Author: Mandy He
-# Date: 23 September 2024
-# Contact: mandyxy.he@mail.utoronto.ca
-# License: N/A
-# Pre-requisites: N/A
-
-
-
-#### Workspace setup ####
-library(opendatatoronto)
-library(tidyverse)
-library(dplyr)
-
-
-#### Download data ####
-
-# get package
-package <- show_package("deaths-of-shelter-residents")
-package
-
-# get all resources for this package
-resources <- list_package_resources("deaths-of-shelter-residents")
-
-# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(resources, tolower(format) %in% c("csv", "geojson"))
-
-# load the Deaths of Shelter Residents data
-Num_death_data <- filter(datastore_resources, row_number() == 1) %>% get_resource()
-
-
-# Deaths of Shelter Residents Mean Age
-Avg_death_age_data <- filter(datastore_resources, row_number() == 2) %>% get_resource()
-
-
-#### Save data ####
-write_csv(Num_death_data, "data/raw_data/Num_death_raw_data.csv")
-write_csv(Avg_death_age_data, "data/raw_data/raw_age_data.csv")

### 2025-06-20 scripts/01-download_data.py

@@ -1,0 +1,46 @@
+#### Preamble ####
+# Purpose: Downloads and saves the data from Open Data Toronto
+# Author: Mandy He
+# Date: 23 September 2024
+# Contact: mandyxy.he@mail.utoronto.ca
+# License: N/A
+# Pre-requisites: N/A
+
+	
+import requests
+from io import StringIO
+import pandas as pd
+ 
+# Toronto Open Data is stored in a CKAN instance. It's APIs are documented here:
+# https://docs.ckan.org/en/latest/api/
+ 
+# To hit our API, you'll be making requests to:
+base_url = "https://ckan0.cf.opendata.inter.prod-toronto.ca"
+ 
+# Datasets are called "packages". Each package can contain many "resources"
+# To retrieve the metadata for this package and its resources, use the package name in this page's URL:
+url = base_url + "/api/3/action/package_show"
+params = { "id": "deaths-of-shelter-residents"}
+package = requests.get(url, params = params).json()
+ 
+dataframe = []
+# To get resource data:
+for idx, resource in enumerate(package["result"]["resources"]):
+ 
+       # for datastore_active resources:
+       if resource["datastore_active"]:
+ 
+           # To get all records in CSV format:
+           url = base_url + "/datastore/dump/" + resource["id"]
+           resource_dump_data = requests.get(url)
+           data = pd.read_csv(StringIO(resource_dump_data.text))
+
+           dataframe.append(data)
+ 
+Num_death_data = dataframe[0]
+Avg_death_age_data = dataframe[1]
+
+
+#### Save data ####
+Num_death_data.to_csv("data/raw_data/Num_death_raw_data.csv", index = False)
+Avg_death_age_data.to_csv("data/raw_data/raw_age_data.csv", index = False)

### 2025-06-20 scripts/02-data_cleaning.R

@@ -1,94 +1,0 @@
-#### Preamble ####
-# Purpose: Cleans the raw data so variable is easier to call
-# Author: Mandy He
-# Date: 23 September 2024
-# Contact: mandyxy.he@mail.utoronto.ca
-# License: N/A
-# Pre-requisites: N/A
-
-#### Workspace setup ####
-library(tidyverse)
-library(dplyr)
-library(tidyr)
-library(readr)
-
-
-#### Clean data ####
-
-## clean the number of death data frist
-raw_death_data <- read_csv("data/raw_data/Num_death_raw_data.csv")
-
-# Display the first few rows of the original data set
-head(raw_death_data)
-
-# Clean the data
-cleaned_death_data <- raw_death_data %>%
-  # Rename columns to remove spaces and special characters
-  rename(
-    id = `_id`,
-    year = `Year`,
-    month = `Month`,
-    total_decedents = `Total decedents`,
-    male = `Male`,
-    female = `Female`,
-    transgender_nonbinary_twospirit = `Transgender/Non-binary/Two-Spirit`
-  ) %>%
-  # Replace "n/a" with 0 in the relevant columns
-  mutate(across(everything(), ~ ifelse(. == "n/a", 0, .))) %>%
-  # Convert relevant columns to numeric
-  mutate(
-    id = as.integer(id),
-    year = as.integer(year),
-    month = as.character(month),
-    total_decedents = as.integer(total_decedents),
-    male = as.integer(male),
-    female = as.integer(female),
-    transgender_nonbinary_twospirit = as.integer(transgender_nonbinary_twospirit)
-  )
-
-# Display the cleaned dataset
-head(cleaned_death_data)
-
-## Clean the Average age of death data.
-raw_age_data <- read_csv("data/raw_data/raw_age_data.csv")
-
-# Display the first few rows of the original dataset
-head(raw_age_data)
-
-# Clean the data
-cleaned_avg_age_death_data <- raw_age_data %>%
-  # Rename columns to remove spaces and special characters
-  rename(
-    id = `_id`,
-    year = `Year`,
-    total_decedents = `Total decedents`,
-    Average_age_of_death_all_decedents = `Average age of death, all decedents`,
-    Male_decedents = `Male decedents`,
-    Average_age_of_death_male_decedents = `Average age of death, male decedents`,
-    Female_decedents = `Female decedents`,
[…]

### 2025-06-20 scripts/02-data_cleaning.py

@@ -1,0 +1,106 @@
+#### Preamble ####
+# Purpose: Cleans the raw data so variable is easier to call
+# Author: Mandy He
+# Date: 23 September 2024
+# Contact: mandyxy.he@mail.utoronto.ca
+# License: N/A
+# Pre-requisites: N/A
+
+#### Workspace setup ####
+import pandas as pd
+from datetime import datetime
+
+#### Clean data ####
+
+## Clean the number of death data first
+raw_death_data = pd.read_csv("data/raw_data/Num_death_raw_data.csv")
+
+# Display the first few rows of the original data set
+print(raw_death_data.head())
+
+# Clean the data
+cleaned_death_data = raw_death_data.rename(columns={
+    "_id": "id",
+    "Year": "year",
+    "Month": "month",
+    "Total decedents": "total_decedents",
+    "Male": "male",
+    "Female": "female",
+    "Transgender/Non-binary/Two-Spirit": "transgender_nonbinary_twospirit"
+})
+
+# Replace "n/a" / null with 0 in all columns
+cleaned_death_data = cleaned_death_data.applymap(
+    lambda x: x if pd.notna(x) else 0
+)
+
+# Convert relevant columns to proper data types
+cleaned_death_data["id"] = cleaned_death_data["id"].astype(int)
+cleaned_death_data["year"] = cleaned_death_data["year"].astype(int)
+cleaned_death_data["month"] = cleaned_death_data["month"].astype(str)
+cleaned_death_data["total_decedents"] = cleaned_death_data["total_decedents"].astype(int)
+cleaned_death_data["male"] = cleaned_death_data["male"].astype(int)
+cleaned_death_data["female"] = cleaned_death_data["female"].astype(int)
+cleaned_death_data["transgender_nonbinary_twospirit"] = cleaned_death_data["transgender_nonbinary_twospirit"].astype(int)
+
+# Display the cleaned dataset
+print(cleaned_death_data.head())
+
+#remove rows after September 2024
+cleaned_death_data = cleaned_death_data[
+    ~(
+        (cleaned_death_data["year"] == 2025) |
+        ((cleaned_death_data["year"] == 2024) & (cleaned_death_data["month"].isin(["Oct", "Nov", "Dec"])))
+    )
+]
+
+## Clean the Average age of death data
+raw_age_data = pd.read_csv("data/raw_data/raw_age_data.csv")
+
+# Display the first few rows of the original dataset
[…]

### 2025-06-20 scripts/03-test_data.py
changed tests

@@ -6,77 +6,63 @@
 # License: N/A
 # Pre-requisites: N/A
 
-
 #### Workspace setup ####
-library(tidyverse)
-# [...UPDATE THIS...]
+import pandas as pd
 
 #### Test data ####
 
 # Test 1: Check for "n/a" Replacement
-test_na_replacement <- function(cleaned_avg_age_death_data) {
-  # Check if any "n/a" values remain in the data
-  has_na <- any(cleaned_avg_age_death_data == "n/a")
+def test_na_replacement(df, df_name):
+    # Check if any 'n/a' strings remain in the data
+    has_na = (df == "n/a").any().any()
 
-  if (!has_na) {
-    print("Test 1 Passed: All 'n/a' values have been replaced with 0.")
-  } else {
-    print("Test 1 Failed: Some 'n/a' values remain in the data.")
-  }
-}
+    if not has_na:
+        print(f"Test 1 Passed: All 'n/a' values have been replaced with 0 in {df_name}.")
+    else:
+        print(f"Test 1 Failed: Some 'n/a' values remain in the data in {df_name}.")
 
 # Run Test 1
-test_na_replacement(cleaned_avg_age_death_data)
-test_na_replacement(cleaned_death_data)
+test_na_replacement(cleaned_avg_age_death_data, "cleaned_avg_age_death_data")
+test_na_replacement(cleaned_death_data, "cleaned_death_data")
 
+# Test 2: Check for Negative Numbers in cleaned_avg_age_death_data
+def test_negative_numbers_avg_age(df):
+    numeric_columns = [
+        "total_decedents",
+        "Average_age_of_death_all_decedents",
+        "Male_decedents",
+        "Average_age_of_death_male_decedents",
+        "Female_decedents",
+        "Average_age_of_death_female_decedents",
+        "Transgender_Non_binary_Two_Spirit_decedents",
+        "Avg_age_of_death_transgender_non_binary_two_spirit_decedents"
+    ]
 
-# Test2: Check for Negative Numbers
-test_negative_numbers <- function(cleaned_avg_age_death_data) {
-  # Define the relevant columns to check for negative numbers
-  numeric_columns <- c(
-    "Total decedents",
-    "Average age of death, all decedents",
-    "Male decedents",
-    "Average age of death, male decedents",
-    "Female decedents",
-    "Average age of death, female decedents",
-    "Transgender/Non-binary/Two-Spirit decedents",
[…]

### 2025-06-20 paper/paper.qmd
changed paper to python

--- 2024-09-27 paper/paper.qmd
@@
 
-```{r}
+```{python}
 #| include: false
@@
 # load necessary package
-library(tidyverse)
-library(ggplot2)
-library(dplyr)
-library(knitr)
-library(tidyr)
-library(gridExtra)
+import pandas as pd
+import matplotlib.pyplot as plt
 
 # load data
-Death_data <- read_csv(here::here("data/analysis_data/Num_death_analysis_data.csv"))
-AvgAge_data <- read_csv(here::here("data/analysis_data/AvgAge_death_analysis_data.csv"))
+Death_data = pd.read_csv("../data/analysis_data/Num_death_analysis_data.csv")
+AvgAge_data = pd.read_csv("../data/analysis_data/AvgAge_death_analysis_data.csv")
 ```
@@
 \newpage
-```{r}
+```{python}
 #| label: tbl-Table1
@@
 # Remove the 'id' column
-AvgAge_data_no_id <- AvgAge_data %>%
-  select(-id)
+AvgAge_data_no_id = AvgAge_data.drop(columns=["id"])
 
-# Display the table without the 'id' column
-kable(AvgAge_data_no_id,
-  col.names = c(
+# Rename columns for display
+AvgAge_data_no_id.columns = [
     "Year", "Total Decedents",
@@
     "Avg Age of Death (Trans/NB/2S)"
+]
+
+# Create the figure and axis
+fig, ax = plt.subplots(figsize=(12, len(AvgAge_data_no_id) * 0.4))  # Adjust height to fit rows
+
+# Hide axis
+ax.axis("off")
+
+# Create table
+table = ax.table(
+    cellText=AvgAge_data_no_id.values,
+    colLabels=AvgAge_data_no_id.columns,
+    loc="center",
+    cellLoc="center"
 )
-)
+
+# Style the table
+table.auto_set_font_size(False)
+table.set_fontsize(8)
+table.scale(1, 1.5)
+
+# Optional title
+plt.title("Table: Variable In The Data – Death Among Shelter Residents (2007–2024)", fontsize=10, pad=20)
+
+plt.tight_layout()
+plt.show()
+
 ```
@@
 
-```{r}
+```{python}
 #| label: fig-TotalNumberofDeathByYear
@@
 
-# creating a line graph
-AvgAge_data |>
-  ggplot(aes(x = year, y = total_decedents)) +
-  geom_point() +
-  geom_line() +
-  labs(title = "Total Shelter Residents Decedents from 2007 to 2024", x = "Year", y = "Total Decedents") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5)) +
-  scale_x_continuous(breaks = unique(AvgAge_data$year)) +
-  scale_y_continuous(breaks = seq(0, max(AvgAge_data$total_decedents), by = 20))
+import matplotlib.pyplot as plt
+
+# Sort by year to ensure correct line order
+AvgAge_data_sorted = AvgAge_data.sort_values("year")
+
+# Create the figure and axis
+plt.figure(figsize=(8, 5))
+plt.plot(AvgAge_data_sorted["year"], AvgAge_data_sorted["total_decedents"], marker="o")
+
+# Set title and labels
+plt.title("Total Shelter Residents Decedents from 2007 to 2024", ha="center")
+plt.xlabel("Year")
+plt.ylabel("Total Decedents")
+
+# Rotate x-axis labels
+plt.xticks(AvgAge_data_sorted["year"], rotation=45)
+
+# Set y-axis ticks at intervals of 20
+plt.yticks(range(0, AvgAge_data_sorted["total_decedents"].max() + 20, 20))
+
+# Apply minimal style (closest match to theme_minimal in ggplot2)
+plt.grid(True, linestyle='--', alpha=0.5)
+plt.tight_layout()
+
+# Show plot (Quarto will capture it)
+plt.show()
+
 ```
@@
 Detailed discussion found in @sec-second-dpoint.
-```{r}
+```{python}
 #| label: fig-TotalDecedentbyMonthcovid
-#| fig-cap: Yearly Mortality rate during COVID (2019-2023)
+#| fig-cap: Yearly Mortality Rate During COVID (2019–2023)
 #| echo: false
+#| warning: false
+#| message: false
 
-# Filter the data for years 2019 to 2022
-Death_data_filtered <- Death_data |>
-  filter(year >= 2019 & year <= 2023)
+import matplotlib.pyplot as plt
+import pandas as pd
 
-# Combine Year and Month into a single column with last two digits of the year
-Death_data_filtered$YearMonth <- paste(substr(Death_data_filtered$year, 3, 4), Death_data_filtered$month, sep = "-")
+# Filter the data for years 2019 to 2023
+Death_data_filtered = Death_data[(Death_data["year"] >= 2019) & (Death_data["year"] <= 2023)].copy()
 
-# Convert YearMonth to a factor with ordered levels
-Death_data_filtered$YearMonth <- factor(Death_data_filtered$YearMonth, levels = unique(Death_data_filtered$YearMonth))
+# Ensure month is ordered and standardized
+month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
+               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
+Death_data_filtered["month"] = pd.Categorical(Death_data_filtered["month"], categories=month_order, ordered=True)
 
-# Plot the graph for filtered data
-Death_data_filtered |>
-  ggplot(aes(x = YearMonth, y = total_decedents, group = 1)) +
-  geom_line(color = "black") +
-  geom_point(color = "black") +
-  scale_x_discrete(breaks = Death_data_filtered$YearMonth[seq(1, nrow(Death_data_filtered), by = 6)]) +
-  labs(
-    title = "Trends in Total Decedents by Month(2019-2023)",
-    x = "Month-Year",
-    y = "Total Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) +
-  theme(plot.title = element_text(hjust = 0.5))
+# Sort by year and month
+Death_data_filtered = Death_data_filtered.sort_values(by=["year", "month"])
+
+# Combine Year and Month into a "YY-MMM" format
+Death_data_filtered["YearMonth"] = Death_data_filtered["year"].astype(str).str[-2:] + "-" + Death_data_filtered["month"].astype(str)
+
+# Plot the data
+plt.figure(figsize=(12, 5))
+plt.plot(Death_data_filtered["YearMonth"], Death_data_filtered["total_decedents"], marker="o", color="black")
+
+# Show x-axis labels every 6 months
+xtick_locs = range(0, len(Death_data_filtered), 6)
+xtick_labels = Death_data_filtered["YearMonth"].iloc[xtick_locs]
+plt.xticks(ticks=xtick_locs, labels=xtick_labels, rotation=90, fontsize=8)
+
+# Labels and title
+plt.title("Yearly Mortality Rate During COVID (2019–2023)", ha="center")
+plt.xlabel("Month-Year")
+plt.ylabel("Total Decedents")
+
+# Minimal grid styling
+plt.grid(True, linestyle="--", alpha=0.5)
+plt.tight_layout()
+
+plt.show()
+
 ```
@@
 These results may be attributed to the proportion of males living in shelters; further details can be found in @sec-3-dpoint.
-```{r}
+```{python}
 #| label: fig-T
@@
 
-# Reshape the data for plotting
-plot_data <- AvgAge_data |>
-  select(year, Male_decedents, Female_decedents) |>
-  pivot_longer(
-    cols = c(Male_decedents, Female_decedents),
-    names_to = "Gender",
-    values_to = "Mortality"
-  )
+import pandas as pd
+import matplotlib.pyplot as plt
+import numpy as np
 
-# Create a new variable for the last two digits of the year with leading zero
-plot_data <- plot_data |>
-  mutate(year_short = sprintf("%02d", year %% 100))
+# Reshape the data for plotting (from wide to long format)
+plot_data = AvgAge_data[["year", "Male_decedents", "Female_decedents"]].copy()
+plot_data = plot_data.melt(id_vars="year", value_vars=["Male_decedents", "Female_decedents"],
+                           var_name="Gender", value_name="Mortality")
 
-# Create the bar graph with adjusted year labels
-plot_data |>
-  ggplot(aes(x = year_short, y = Mortality, fill = Gender)) +
-  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
-  labs(
-    title = "Mortality Numbers of Male Vs Female Decedents over 2007 to 2024",
-    x = "Year (2000s)",
-    y = "Number of Decedents"
-  ) +
-  scale_fill_manual(values = c("Male_decedents" = "blue", "Female_decedents" = "pink")) +
-  theme_minimal() +
-  theme(
-    legend.title = element_blank(),
-    legend.position = "bottom"
-  ) +
-  theme(plot.title = element_text(hjust = 0.5))
+# Create a new variable for the last two digits of the year, with leading zero
+plot_data["year_short"] = plot_data["year"] % 100
+plot_data["year_short"] = plot_data["year_short"].apply(lambda x: f"{x:02d}")
+
+# Create a grouped bar chart
+years = plot_data["year_short"].unique()
+x = np.arange(len(years))  # the label locations
+
+male_values = plot_data[plot_data["Gender"] == "Male_decedents"]["Mortality"].values
+female_values = plot_data[plot_data["Gender"] == "Female_decedents"]["Mortality"].values
+
+width = 0.35  # width of the bars
+
+fig, ax = plt.subplots(figsize=(10, 5))
+ax.bar(x - width/2, male_values, width, label="Male", color="blue")
+ax.bar(x + width/2, female_values, width, label="Female", color="pink")
+
+# Labels and formatting
+ax.set_xlabel("Year (2000s)")
+ax.set_ylabel("Number of Decedents")
+ax.set_title("Mortality Numbers of Male Vs Female Decedents over 2007 to 2024", pad=15)
+ax.set_xticks(x)
+ax.set_xticklabels(years, rotation=0)
+ax.legend(title="", loc="lower center", bbox_to_anchor=(0.5, -0.2), ncol=2)
+ax.grid(True, axis='y', linestyle='--', alpha=0.5)
+
+plt.tight_layout()
+plt.show()
+
 ```
@@
 
-```{r}
+```{python}
 #| label: fig-AvgAgeofDecedentsFemaleVSMale
-#| fig-cap: Average Age of Death (Female VS Male)
+#| fig-cap: Average Age of Death (Female VS Male) 2007–2024
 #| echo: false
@@
 
-# Reshape the data for plotting average age of death
-avg_age_data <- AvgAge_data |>
-  select(year, Average_age_of_death_male_decedents, Average_age_of_death_female_decedents) %>%
-  pivot_longer(
-    cols = c(Average_age_of_death_male_decedents, Average_age_of_death_female_decedents),
-    names_to = "Gender",
-    values_to = "Average_Age"
-  ) |>
-  mutate(year_short = sprintf("%02d", year %% 100))
+import pandas as pd
+import matplotlib.pyplot as plt
 
-# Create the line graph for average age of death
-avg_age_data |>
-  ggplot(aes(x = year_short, y = Average_Age, color = Gender, group = Gender)) +
-  geom_line(size = 1) +
-  geom_point(size = 2) +
-  labs(
-    title = "Average Age of Death for Male Vs Female Decedents",
-    x = "Year",
-    y = "Average Age of Death"
-  ) +
-  scale_color_manual(values = c(
-    "Average_age_of_death_male_decedents" = "blue",
-    "Average_age_of_death_female_decedents" = "pink"
-  )) +
-  theme_minimal() +
-  theme(
-    legend.title = element_blank(),
-    legend.position = "bottom"
-  ) +
-  theme(plot.title = element_text(hjust = 0.5))
+# Reshape the data from wide to long format
+avg_age_data = AvgAge_data[["year", "Average_age_of_death_male_decedents", "Average_age_of_death_female_decedents"]].copy()
+avg_age_data = avg_age_data.melt(
+    id_vars="year",
+    value_vars=["Average_age_of_death_male_decedents", "Average_age_of_death_female_decedents"],
+    var_name="Gender",
+    value_name="Average_Age"
+)
+
+# Create new column with last two digits of the year
+avg_age_data["year_short"] = avg_age_data["year"] % 100
+avg_age_data["year_short"] = avg_age_data["year_short"].apply(lambda x: f"{x:02d}")
+
+# Pivot back to format lines cleanly
+pivoted = avg_age_data.pivot(index="year_short", columns="Gender", values="Average_Age")
+
+# Plot
+plt.figure(figsize=(10, 5))
+plt.plot(pivoted.index, pivoted["Average_age_of_death_male_decedents"], marker="o", label="Male", color="blue")
+plt.plot(pivoted.index, pivoted["Average_age_of_death_female_decedents"], marker="o", label="Female", color="pink")
+
+# Titles and labels
+plt.title("Average Age of Death for Male Vs Female Decedents", ha="center")
+plt.xlabel("Year (in 2000s)")
+plt.ylabel("Average Age of Death")
+
+# Formatting
+plt.xticks(rotation=0)
+plt.legend(title="", loc="lower center", bbox_to_anchor=(0.5, -0.25), ncol=2)
+plt.grid(True, linestyle="--", alpha=0.5)
+plt.tight_layout()
+
+plt.show()
 ```
@@
 
-```{r}
+```{python}
 #| label: fig-2007-2010
-#| fig-cap: Monthly Mortality rate from 2007-2010
+#| fig-cap: Monthly Mortality Rate from 2007–2010
 #| echo: false
@@
 
-# make sure that the month is in order from Jan to Dec.
-Death_data$month <- factor(Death_data$month, levels = month.abb)
+import pandas as pd
+import matplotlib.pyplot as plt
 
-# Filter data for the years 2007 and 2008
-current_year_data_2007 <- Death_data %>% filter(year == 2007)
-current_year_data_2008 <- Death_data %>% filter(year == 2008)
-current_year_data_2009 <- Death_data %>% filter(year == 2009)
-current_year_data_2010 <- Death_data %>% filter(year == 2010)
+# Ensure month column is a categorical with correct month order
+month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
+               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
+Death_data["month"] = pd.Categorical(Death_data["month"], categories=month_order, ordered=True)
 
-# Create the bar plots for 2007 and 2008
-plot_2007 <- ggplot(current_year_data_2007, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2007",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
+# Filter for each year
+data_2007 = Death_data[Death_data["year"] == 2007]
+data_2008 = Death_data[Death_data["year"] == 2008]
+data_2009 = Death_data[Death_data["year"] == 2009]
+data_2010 = Death_data[Death_data["year"] == 2010]
 
-plot_2008 <- ggplot(current_year_data_2008, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2008",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
+# Set up subplot layout
+fig, axs = plt.subplots(2, 2, figsize=(12, 8))
+axs = axs.flatten()  # Flatten to 1D for easy indexing
 
-plot_2009 <- ggplot(current_year_data_2009, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2009",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
+# Year-wise loop for consistent plotting
+for i, (data, year) in enumerate(zip([data_2007, data_2008, data_2009, data_2010], [2007, 2008, 2009, 2010])):
+    axs[i].bar(data["month"], data["total_decedents"], color="black")
+    axs[i].set_title(f"Total Decedents by Month in {year}", fontsize=10, ha="center")
+    axs[i].set_xlabel("Month")
+    axs[i].set_ylabel("# Decedents")
+    axs[i].tick_params(axis="x", rotation=45)
+    axs[i].grid(True, axis="y", linestyle="--", alpha=0.4)
 
-plot_2010 <- ggplot(current_year_data_2010, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2010",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
-
-# Arrange plots side by side
-grid.arrange(plot_2007, plot_2008, plot_2009, plot_2010, ncol = 2)
+# Layout adjustment
+plt.tight_layout()
+plt.show()
 ```
 
-```{r}
+```{python}
 #| label: fig-2011-2014
-#| fig-cap: Monthly Mortality rate from 2011-2014
+#| fig-cap: Monthly Mortality Rate from 2011–2014
 #| echo: false
@@
 
-# make sure that the month is in order from Jan to Dec.
-Death_data$month <- factor(Death_data$month, levels = month.abb)
+import matplotlib.pyplot as plt
+import pandas as pd
 
-# Filter data for the years
-current_year_data_2011 <- Death_data %>% filter(year == 2011)
-current_year_data_2012 <- Death_data %>% filter(year == 2012)
-current_year_data_2013 <- Death_data %>% filter(year == 2013)
-current_year_data_2014 <- Death_data %>% filter(year == 2014)
+# Ensure months are in correct order
+month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
+               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
+Death_data["month"] = pd.Categorical(Death_data["month"], categories=month_order, ordered=True)
 
-# Create the bar plots
-plot_2011 <- ggplot(current_year_data_2011, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2011",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
+# Filter data by year
+data_2011 = Death_data[Death_data["year"] == 2011]
+data_2012 = Death_data[Death_data["year"] == 2012]
+data_2013 = Death_data[Death_data["year"] == 2013]
+data_2014 = Death_data[Death_data["year"] == 2014]
 
-plot_2012 <- ggplot(current_year_data_2012, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2012",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
+# Set up 2x2 subplot layout
+fig, axs = plt.subplots(2, 2, figsize=(12, 8))
+axs = axs.flatten()
 
-plot_2013 <- ggplot(current_year_data_2013, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2013",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
+# Plot each year
+for i, (data, year) in enumerate(zip([data_2011, data_2012, data_2013, data_2014], [2011, 2012, 2013, 2014])):
+    axs[i].bar(data["month"], data["total_decedents"], color="black")
+    axs[i].set_title(f"Total Decedents by Month in {year}", fontsize=10, ha="center")
+    axs[i].set_xlabel("Month")
+    axs[i].set_ylabel("# Decedents")
+    axs[i].tick_params(axis="x", rotation=45)
+    axs[i].grid(True, axis="y", linestyle="--", alpha=0.4)
 
-plot_2014 <- ggplot(current_year_data_2014, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2014",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
-
-# Arrange plots side by side
-grid.arrange(plot_2011, plot_2012, plot_2013, plot_2014, ncol = 2)
+# Adjust layout for better spacing
+plt.tight_layout()
+plt.show()
 ```
 
-```{r}
+```{python}
 #| label: fig-2015-2018
-#| fig-cap: Monthly Mortality rate from 2015-2018
+#| fig-cap: Monthly Mortality Rate from 2015–2018
 #| echo: false
@@
 
-# make sure that the month is in order from Jan to Dec.
-Death_data$month <- factor(Death_data$month, levels = month.abb)
+import matplotlib.pyplot as plt
+import pandas as pd
 
-# Filter data for the years
+# Ensure months are in correct order
+month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
+               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
+Death_data["month"] = pd.Categorical(Death_data["month"], categories=month_order, ordered=True)
 
-current_year_data_2015 <- Death_data %>% filter(year == 2015)
-current_year_data_2016 <- Death_data %>% filter(year == 2016)
-current_year_data_2017 <- Death_data %>% filter(year == 2017)
-current_year_data_2018 <- Death_data %>% filter(year == 2018)
+# Filter data by year
+data_2015 = Death_data[Death_data["year"] == 2015]
+data_2016 = Death_data[Death_data["year"] == 2016]
+data_2017 = Death_data[Death_data["year"] == 2017]
+data_2018 = Death_data[Death_data["year"] == 2018]
 
-# Create the bar plots
+# Set up 2x2 subplot layout
+fig, axs = plt.subplots(2, 2, figsize=(12, 8))
+axs = axs.flatten()
 
-plot_2015 <- ggplot(current_year_data_2015, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2015",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
+# Plot each year's data
+for i, (data, year) in enumerate(zip([data_2015, data_2016, data_2017, data_2018], [2015, 2016, 2017, 2018])):
+    axs[i].bar(data["month"], data["total_decedents"], color="black")
+    axs[i].set_title(f"Total Decedents by Month in {year}", fontsize=10, ha="center")
+    axs[i].set_xlabel("Month")
+    axs[i].set_ylabel("# Decedents")
+    axs[i].tick_params(axis="x", rotation=45)
+    axs[i].grid(True, axis="y", linestyle="--", alpha=0.4)
 
-plot_2016 <- ggplot(current_year_data_2016, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2016",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
+# Adjust layout
+plt.tight_layout()
+plt.show()
 
-plot_2017 <- ggplot(current_year_data_2017, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2017",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
-
-plot_2018 <- ggplot(current_year_data_2018, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2018",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
-
-# Arrange plots side by side
-grid.arrange(plot_2015, plot_2016, plot_2017, plot_2018, ncol = 2)
 ```
 
-```{r}
+```{python}
 #| label: fig-2019-2022
-#| fig-cap: Monthly Mortality rate from 2019-2022
+#| fig-cap: Monthly Mortality Rate from 2019–2022
 #| echo: false
@@
 
-# make sure that the month is in order from Jan to Dec.
-Death_data$month <- factor(Death_data$month, levels = month.abb)
+import matplotlib.pyplot as plt
+import pandas as pd
 
-# Filter data for the years
+# Ensure months are in correct order
+month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
+               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
+Death_data["month"] = pd.Categorical(Death_data["month"], categories=month_order, ordered=True)
 
-current_year_data_2019 <- Death_data %>% filter(year == 2019)
-current_year_data_2020 <- Death_data %>% filter(year == 2020)
-current_year_data_2021 <- Death_data %>% filter(year == 2021)
-current_year_data_2022 <- Death_data %>% filter(year == 2022)
+# Filter data by year
+data_2019 = Death_data[Death_data["year"] == 2019]
+data_2020 = Death_data[Death_data["year"] == 2020]
+data_2021 = Death_data[Death_data["year"] == 2021]
+data_2022 = Death_data[Death_data["year"] == 2022]
 
-# Create the bar plots
-plot_2019 <- ggplot(current_year_data_2019, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2019",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
+# Set up 2x2 subplot layout
+fig, axs = plt.subplots(2, 2, figsize=(12, 8))
+axs = axs.flatten()
 
-plot_2020 <- ggplot(current_year_data_2020, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2020",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
+# Plot each year
+for i, (data, year) in enumerate(zip([data_2019, data_2020, data_2021, data_2022], [2019, 2020, 2021, 2022])):
+    axs[i].bar(data["month"], data["total_decedents"], color="black")
+    axs[i].set_title(f"Total Decedents by Month in {year}", fontsize=10, ha="center")
+    axs[i].set_xlabel("Month")
+    axs[i].set_ylabel("# Decedents")
+    axs[i].tick_params(axis="x", rotation=45)
+    axs[i].grid(True, axis="y", linestyle="--", alpha=0.4)
 
-plot_2021 <- ggplot(current_year_data_2021, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2021",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
+# Adjust layout
+plt.tight_layout()
+plt.show()
 
-plot_2022 <- ggplot(current_year_data_2022, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2022",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
-
-# Arrange plots side by side
-grid.arrange(plot_2019, plot_2020, plot_2021, plot_2022, ncol = 2)
 ```
 
-```{r}
+```{python}
 #| label: fig-2023-2024
-#| fig-cap: Monthly Mortality rate from 2023-2024
+#| fig-cap: Monthly Mortality Rate from 2023–2024
 #| echo: false
@@
 
-# make sure that the month is in order from Jan to Dec.
-Death_data$month <- factor(Death_data$month, levels = month.abb)
+import matplotlib.pyplot as plt
+import pandas as pd
 
-# Filter data for the years
-current_year_data_2023 <- Death_data %>% filter(year == 2023)
-current_year_data_2024 <- Death_data %>% filter(year == 2024)
+# Ensure months are ordered from Jan to Dec
+month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
+               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
+Death_data["month"] = pd.Categorical(Death_data["month"], categories=month_order, ordered=True)
 
-# Create the bar plots
-plot_2023 <- ggplot(current_year_data_2023, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2023",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
+# Filter data for 2023 and 2024
+data_2023 = Death_data[Death_data["year"] == 2023]
+data_2024 = Death_data[Death_data["year"] == 2024]
 
-plot_2024 <- ggplot(current_year_data_2024, aes(x = month, y = total_decedents)) +
-  geom_bar(stat = "identity", fill = "black") +
-  labs(
-    title = "Total Decedents by Month in 2024",
-    x = "Month",
-    y = "# Decedents"
-  ) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45)) +
-  theme(plot.title = element_text(hjust = 0.5, size = 10))
+# Set up 1x2 subplot layout
+fig, axs = plt.subplots(1, 2, figsize=(12, 4))
 
-# Arrange plots side by side
-grid.arrange(plot_2023, plot_2024, ncol = 2)
+# Plot each year
+for ax, data, year in zip(axs, [data_2023, data_2024], [2023, 2024]):
+    ax.bar(data["month"], data["total_decedents"], color="black")
+    ax.set_title(f"Total Decedents by Month in {year}", fontsize=10, ha="center")
+    ax.set_xlabel("Month")
+    ax.set_ylabel("# Decedents")
+    ax.tick_params(axis="x", rotation=45)
+    ax.grid(True, axis="y", linestyle="--", alpha=0.4)
+
+# Adjust layout
+plt.tight_layout()
+plt.show()
+
 ```

### 2025-06-20 scripts/02-data_cleaning.py

@@ -46,11 +46,11 @@ cleaned_death_data["transgender_nonbinary_twospirit"] = cleaned_death_data["tran
 # Display the cleaned dataset
 print(cleaned_death_data.head())
 
-#remove rows after September 2024
+#remove rows after & including September 2024
 cleaned_death_data = cleaned_death_data[
     ~(
         (cleaned_death_data["year"] == 2025) |
-        ((cleaned_death_data["year"] == 2024) & (cleaned_death_data["month"].isin(["Oct", "Nov", "Dec"])))
+        ((cleaned_death_data["year"] == 2024) & (cleaned_death_data["month"].isin(["Sep", "Oct", "Nov", "Dec"])))
     )
 ]
 

### 2025-06-24 paper/paper.qmd
anonymized, changed citations, chnaged title, still issue with cross ref

@@
 ---
-title: "Mortality Trend Among Shelter Residents: "
-subtitle: "Closely Tied To Virus Outbreaks"
+title: "Analyzing Mortality Among Shelter Residents"
+subtitle: "Evidence of Strong Links to Viral Outbreaks"
 author: 
-  - Mandy He
-thanks: "Code and data are available at: https://github.com/MandyHe7/Mortality-Trend-Among-Shelter-Residents"
+  - Vivian Guo
+thanks: "Code and data are available at: https://github.com/EveHughes/Mortality-Trend-Among-Shelter-Residents"
 date: Septmeber 27 2024
@@
 Thus, examining these patterns can help improve provisions and reduce the number of deaths among both this vulnerable population and society at large."
-format: pdf
+format:
+  pdf:
+    pdf-engine: lualatex
 number-sections: true
@@
 The data used in this paper is access from Open Data Toronto and the particular data set used was the Deaths of Shelter Residents (@toronto_deaths_shelter_residents).
-To analysis the data and creating graphs using the data, following package that was build in the (R program @citeR) was used: tidyverse (@tidyverse), dplyr (@dplyr), tidyr (@tidyr), gridExtra (@gridExtra), knitr (@knitr), and ggplot2 (@ggplot2).
-LLM (@openai2023chatgpt) was also use in term of helping with coding in R to graph data, simulating data, cleaning data, and revising writing.
+To analysis the data and creating graphs using the data, following package that was build in the (Python program @python) was used: pandas (@pandas), numpy (@numpy), and matplotlib (@matplotlib).
 

### 2025-06-24 scripts/00-simulate_data.py

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Simulates a sample data of Deaths Among Shelter Residents from 2007 to 2024
-# Author: Mandy He
+# Author: Vivian Guo
 # Date: 23 September 2024
-# Contact: mandyxy.he@mail.utoronto.ca
+# Contact: vivian.guo@mail.utoronto.ca
 # License: N/A
 # Pre-requisites: N/A
 

### 2025-06-24 scripts/01-download_data.py

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Downloads and saves the data from Open Data Toronto
-# Author: Mandy He
+# Author: Vivian Guo
 # Date: 23 September 2024
-# Contact: mandyxy.he@mail.utoronto.ca
+# Contact: vivian.guo@mail.utoronto.ca
 # License: N/A
 # Pre-requisites: N/A
 

### 2025-06-24 scripts/02-data_cleaning.py

@@ -1,14 +1,13 @@
 #### Preamble ####
 # Purpose: Cleans the raw data so variable is easier to call
-# Author: Mandy He
+# Author: Vivian Guo
 # Date: 23 September 2024
-# Contact: mandyxy.he@mail.utoronto.ca
+# Contact: vivian.guo@mail.utoronto.ca
 # License: N/A
 # Pre-requisites: N/A
 
 #### Workspace setup ####
 import pandas as pd
-from datetime import datetime
 
 #### Clean data ####
 

### 2025-06-24 scripts/03-test_data.py

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Tests for "n/a" Replacement and Negative Numbers.
-# Author: Mandy He
+# Author: Vivian Guo
 # Date: 23 September 2024
-# Contact: mandyxy.he@mail.utoronto.ca
+# Contact: vivian.guo@mail.utoronto.ca
 # License: N/A
 # Pre-requisites: N/A
 

### 2025-06-24 scripts/04-Code_auto_format.R

@@ -1,9 +1,0 @@
-# Load the styler package
-library(styler)
-
-# Define files to format
-files_to_format <- c("scripts/00-simulate_data.R", "scripts/01-download_data.R", "scripts/02-data_cleaning.R",
-                     "scripts/03-test_data.R", "paper/paper.qmd")
-
-# Loop through each file and format it
-lapply(files_to_format, style_file)
